---
  title: "Fetal origins of malignant germ cell tumours"
author: "Thomas R W Oliver"
output: 
  html_document:
  toc: true
toc_depth: 2
number_sections: true
editor_options:
  chunk_output_type: console
---
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

If you have any queries or have identified potential errors, please email to3\\@sanger.ac.uk.

#Bespoke DNA filtering and processing steps

These were necessary to remove sequencing and mapping artefacts that passed the standard low input pipeline variant calling. Usually, these would be removed by fitting a beta-binomial distribution and assessing for overdispersion of VAFs across all samples taken from a patient. Here, because most cases only had a single matched normal and the remaining tumour samples shared many variants at similar VAFs, this approach would remove true variants.

##Site specific error rate estimation

Adapted from Coorens et al 2019.

###SNV filtering

```{r}

#-------------------------------------------------
# Libraries
#-------------------------------------------------

library("GenomicRanges")
library("deepSNV")
library("Rsamtools")

logbb = deepSNV:::logbb
dbetabinom = VGAM::dbetabinom

#-------------------------------------------------
# Functions
#-------------------------------------------------

estimateRho_gridml = function(x, mu) {
  # Estimate rho by MLE grid approach
  rhovec = 10^seq(-6,-0.5,by=0.05) # rho will be bounded within 1e-6 and 0.32
  mm = x[,2]
  cov = c(x[,1])
  ll = sapply(rhovec, function(rhoj) sum(dbetabinom(x=mm, size=cov, rho=rhoj, prob=mu, log=T)))
  rhovec[ll==max(ll)][1]
}

shearwater_probability=function(patient, save=NULL, path_prefix='', rho=10^-3){
  #Function to calculate probability of presence of mutation based on Shearwarer
  #'patient' is the name of the patient-specific subfolder
  #'path_prefix' is any prefix to the path necessary to find that subfolder
  #'rho' is the constant for the overdispersion parameter. If rho=NULL, calculate
  # it from the data (much slower)
  #'save' is path for output. If NULL, returns matrix
  # output of this function will be a matrix (muts by samples) of p values
  
  
  #A file of mutations in patient subdirectory (format: Chr_Ref_Pos_Alt)
  Muts_patient = read.table(paste0(path_prefix,patient,"/All_mutations_filtered.txt"))[,1] 
  #A file of with the sample names belonging to this patient
  case_samples=read.table(paste0(path_prefix,patient,"/samples.txt"))[,1]
  normal_panel = normal_samples
  norm_all_counts = all_counts[normal_panel,,]
  coords_proj = substr(Muts_patient,1,nchar(Muts_patient)-4)
  case_all_counts = all_counts[case_samples,,]
 
  avg_depth_list = c()
  for (i in 1:length(coords_proj)){
    mean_d = sum(case_all_counts[,i,]) / length(case_samples)
    avg_depth_list = c(avg_depth_list, mean_d)
  }
  
  Muts_patient = Muts_patient[avg_depth_list > 10] #Removes remaining globally low coverage variants which were the result of mapping issues
  coords_proj = coords_proj[avg_depth_list > 10]

  #Set up pval matrix
  pval_mat = matrix(1,ncol=length(case_samples),nrow=length(Muts_patient))
  rownames(pval_mat)=Muts_patient
  colnames(pval_mat)=case_samples
  
  
  Alt=substr(Muts_patient,nchar(Muts_patient),nchar(Muts_patient))
  Ref=substr(Muts_patient,nchar(Muts_patient)-2,nchar(Muts_patient)-2)
  
  for (s in case_samples){
    rho_est=rep(NA,length(coords_proj))
    test_counts = all_counts[s,coords_proj,]
    for (k in 1:length(coords_proj)) {
      n = sum(test_counts[coords_proj[k],])
      x = test_counts[coords_proj[k],Alt[k]]
      
      N_indiv = rowSums(norm_all_counts[,coords_proj[k],])
      X_indiv = norm_all_counts[,coords_proj[k],c("A","C","G","T")!=Ref[k]]
      pseudo = .Machine$double.eps    
      N=sum(N_indiv)
      X=sum(X_indiv)
      
      mu = max(X,pseudo)/max(N,pseudo)
      counts = cbind(N,X)
      if(is.null(rho)) rho = estimateRho_gridml(counts,mu)
      rdisp = (1 - rho)/rho
      
      prob0 = (X + x)/(N + n); prob0[prob0==0] = pseudo; prob0[prob0==1] = 1 - pseudo
      prob1s = x/(n + pseudo); prob1s[prob1s==0] = pseudo; prob1s[prob1s==1] = 1 - pseudo
      prob1c = X/(N + pseudo); prob1c[prob1c==0] = pseudo; prob1c[prob1c==1] = 1 - pseudo
      
      prob1s = pmax(prob1s,prob1c) # Min error rate is that of the population (one-sided test)
      nu0 = prob0 * rdisp; nu1s = prob1s * rdisp; nu1c = prob1c * rdisp; 
      
      # Likelihood-Ratio Tests
      LL = logbb(x, n, nu0, rdisp) + logbb(X, N, nu0, rdisp) - logbb(x, n, nu1s, rdisp) - logbb(X, N, nu1c, rdisp)
      pvals = pchisq(-2*LL, df=1, lower.tail=F)/2 # We divide by 2 as we are performing a 1-sided test
      # Saving the result
      pval_mat[k,s] = pvals
    } 
  }
  if(is.null(save)){
    return(pval_mat)
  }else{
    write.table(pval_mat,save)
  }
}

#-------------------------------------------------
# Input
#-------------------------------------------------

#setwd('/lustre/scratch117/casm/team294/to3/testes/tumour/allelecount/20200511')
#options(stringsAsFactors = F)

# Vector of normal reference samples, taken from a separate cohort of patients
normal_samples = read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/reference_datasets/reference_lcm_normal_panel.txt")[,1] 

# Vector of all samples, including the reference panel and samples of interest
tumour_samples = read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/reference_datasets/gct_samples.txt")[,1]
samples <- c(tumour_samples, normal_samples)

# "Bed" file of all mutations to be considered (across all patients)
# Format: Chr Ref Pos Alt
muts = read.table("GCT_SSM.snvs.bed") 
coords = paste(muts$V1, muts$V2, sep="_")

# Read in data from AlleleCounter
all_counts = array(0,dim=c(length(samples), length(coords), 4),
                   dimnames = list(samples, coords, c("A","C","G","T")))

print(length(samples)) #381 samples, 250 reference samples, 131 GCT genomes

for (k in 1:length(samples)){
  #Read in allele counts per sample
  if(file.exists(paste0("output/", samples[k],".txt"))){
    data=read.table(paste0("output/", samples[k],".txt"),comment.char = '',header=T)
    muts_data=paste(data$X.CHR,data$POS,sep="_")
    data=data[!duplicated(muts_data),]
    muts_data=muts_data[!duplicated(muts_data)]
    rownames(data)=muts_data
    all_counts[k,,]=as.matrix(data[coords,3:6])
  }
  print(k)
}

#create list of study patients
#setwd('/lustre/scratch117/casm/team294/to3/testes/tumour/cgpVAF_phase2')
sample.list <- read.table('patients_all.txt', header = F)[,1]

#generate mutations per patient list
patient.list <- unique(substr(sample.list, 0, 7))

for(i in patient.list){
  data <- read.table(paste0(i, '/', list.files(paste0(i, '/'), pattern = '_snp_vaf_nohash.tsv')[1]), comment.char = "", header = T)
  Muts = paste(data$Chrom, data$Pos, data$Ref, data$Alt, sep="_")
  write.table(Muts, paste0(i, '/All_mutations_filtered.txt'), col.names = F, row.names = F, quote = F)
}

#generate a list of samples per tumour
for(i in patient.list){
  write.table(unlist(strsplit(list.files(paste0(i, '/'), pattern = '_snps_geno.final.vcf'), '_snps_geno.final.vcf')), paste0(i, '/samples.txt'), col.names = F, row.names = F, quote = F)
}

# read in the list of study patients
patients = read.table("patients_all.txt")[,1]

#-------------------------------------------------
# Run
#-------------------------------------------------

for (patient in patients){
  shearwater_probability(patient=patient, save=paste0(patient, "/shearwater_snv_pval_mat.txt"))
  print(patient)
}

#get an idea of the mutation burden per sample after this filtering step
post_swater_mb <- c()
for (patient in patients){
  pval_mat = read.table(paste0(patient, "/shearwater_snv_pval_mat.txt"), row.names = 1, sep = ' ', header = T)
  qval_mat = apply(pval_mat,2,function(x) p.adjust(x,method="BH", n = length(as.matrix(pval_mat))))
  post_swater_mb <- c(post_swater_mb, colSums(qval_mat < 0.001, na.rm = T))
  }

#BH correction for multiple testing
for (patient in patients){
  pval_mat = read.table(paste0(patient, "/shearwater_snv_pval_mat.txt"), row.names = 1, sep = ' ', header = T)
  qval_mat = apply(pval_mat,2,function(x) p.adjust(x,method="BH", n = length(as.matrix(pval_mat))))
  write.table(qval_mat, paste0(patient, "/shearwater_snv_qval_mat.txt"), row.names = T, sep = '\t', quote = F)
  }

#filter vcf to leave only variants that pass the site specific error threshold

library(VariantAnnotation)

for (patient in patients){
  qval_mat = read.table(paste0(patient, "/shearwater_snv_qval_mat.txt"), row.names = 1, sep = '\t', header = T)
  samples = read.table(paste0(patient, "/samples.txt"), sep = '\t', header = F)[,1]
  for (sample in samples){
    data = readVcf(paste0(patient, '/', sample, '_snps_geno.final.vcf'))
    reads = cbind((geno(data)$FAZ)[,2]+(geno(data)$RAZ)[,2],(geno(data)$FCZ)[,2]+(geno(data)$RCZ)[,2],
                (geno(data)$FGZ)[,2]+(geno(data)$RGZ)[,2],(geno(data)$FTZ)[,2]+(geno(data)$RTZ)[,2])
    Var = data.frame(Chr=as.character(seqnames(rowRanges(data))),
                   Pos=start(ranges(rowRanges(data))),
                   Ref=as.character(ref(data)))
    Alt_tmp = CharacterList(alt(data))
    Var$Alt = as.character(unlist(Alt_tmp))
    Var$NR=rowSums(reads)
    Var$NV=NA
    colnames(reads)=c("A","C","G","T")
    for (k in c("A","C","G","T")){
      Var$NV[Var$Alt==k] = reads[Var$Alt==k,k]
    }
    Var$VAF=Var$NV/Var$NR
    Var$ID = paste(Var$Chr, Var$Pos, Var$Ref, Var$Alt, sep = '_')
    qval_patient_sig_muts = row.names(qval_mat[qval_mat[,sample] < 0.001,])
    swater_filtered_muts = Var[Var$ID %in% qval_patient_sig_muts,]
    write.table(swater_filtered_muts, paste0(patient, '/', sample, '_post_swater_final_snvs.txt'), sep = '\t', quote = F, row.names = F)
  }
}

```

###Indel filtering

Similar to above but, rather than using allelecounter, we used cgpVAF/exonerate for the pileup. It was also run on a per patient basis rather than a single pileup across the cohort.

```{r}

#-------------------------------------------------
# Libraries
#-------------------------------------------------

library("GenomicRanges")
library("deepSNV")
library("Rsamtools")

logbb = deepSNV:::logbb
dbetabinom = VGAM::dbetabinom

#-------------------------------------------------
# Functions
#-------------------------------------------------

estimateRho_gridml = function(x, mu) {
  # Estimate rho by MLE grid approach
  rhovec = 10^seq(-6,-0.5,by=0.05) # rho will be bounded within 1e-6 and 0.32
  mm = x[,2]
  cov = c(x[,1])
  ll = sapply(rhovec, function(rhoj) sum(dbetabinom(x=mm, size=cov, rho=rhoj, prob=mu, log=T)))
  rhovec[ll==max(ll)][1]
}

shearwater_indel_probability=function(patient, save=NULL, path_prefix='', rho=10^-3){
  #Function to calculate probability of presence of mutation based on Shearwater
  #'patient' is the name of the patient-specific subfolder
  #'path_prefix' is any prefix to the path necessary to find that subfolder
  #'rho' is the constant for the overdispersion parameter. If rho=NULL, calculate
  # it from the data (much slower)
  #'save' is path for output. If NULL, returns matrix
  # output of this function will be a matrix (muts by samples) of p values
  
  Muts_patient = read.table(paste0(path_prefix, patient, "/All_mutations_filtered.txt"))[,1] #A file of mutations in patient subdirectory (format: Chr_Ref_Pos_Alt)
  Muts_patient = unique(Muts_patient) #to avoid duplicating variants
  case_samples = read.table(paste0(path_prefix, patient,"/samples.txt"))[,1] #A file of with the sample names belonging to this patient
  NR = read.table(paste0(path_prefix, 'NR_all_samples_all_muts.txt'), row.names = 1, header = T, sep = '\t') #total read depth at each variant locus called in this patient across all samples
  NR[NR == 0] = 1 #to prevent NAs
  NV = read.table(paste0(path_prefix, 'NV_all_samples_all_muts.txt'), row.names = 1, header = T, sep = '\t') #variant read depth per variant per sample in patient
  matched_normals = read.table(paste0(path_prefix, 'matched_normals.txt'), header = F, sep = '\t')[,1]
  matched_normal = matched_normals[grepl(patient, matched_normals)] #picks only this patient's matched normal
  
  NR = NR[Muts_patient,] #only mutations relevant to patient
  NV = NV[Muts_patient,]
  
  Depth_filter = rowMeans(NR[, case_samples]) > 10 #Removes remaining globally low coverage variants which were the result of mapping issues
  
  NR = NR[Depth_filter,]
  NV = NV[Depth_filter,]
  
  Min_var_depth = rowSums(NV[, case_samples] >= 5) > 0 #removes low VAF sequencing artefacts. This ensures only variants called in a sample to at least 5 variant reads are kept.
  
  NR = NR[Min_var_depth,]
  NV = NV[Min_var_depth,]
  
  Muts_patient = row.names(NR) #new, slimmer list to run shearwater-like approach on
  
  #Select the normal panel not belonging to this patient with the pileup of sites where variants called in patient's tumour samples
  normal_panel = normal_samples
  norm_NR = NR[, normal_panel]
  norm_NV = NV[, normal_panel]
  
  #Set up pval matrix
  pval_mat = matrix(1,ncol=length(case_samples),nrow=length(Muts_patient))
  rownames(pval_mat)=Muts_patient
  colnames(pval_mat)=case_samples
  
  coords_proj = paste(matrix(unlist(strsplit(Muts_patient, '_')), ncol = 4, byrow = T)[,1], matrix(unlist(strsplit(Muts_patient, '_')), ncol = 4, byrow = T)[,2], sep = '_')
  
  Alt = matrix(unlist(strsplit(Muts_patient, '_')), ncol = 4, byrow = T)[,4]
  Ref = matrix(unlist(strsplit(Muts_patient, '_')), ncol = 4, byrow = T)[,3]
  
  for (s in case_samples){
    rho_est = rep(NA,length(coords_proj))
    test_counts_NR = NR[, s]
    test_counts_NV = NV[, s]
    
    for (k in 1:length(coords_proj)) {
      n = test_counts_NR[k]
      x = test_counts_NV[k]
      N_indiv = norm_NR[k,]
      X_indiv = norm_NV[k,]
      pseudo = .Machine$double.eps    
      N = sum(N_indiv)
      X = sum(X_indiv)
      mu = max(X,pseudo)/max(N,pseudo)
      counts = cbind(N,X)
      
      if(is.null(rho)) rho = estimateRho_gridml(counts,mu)
      rdisp = (1 - rho)/rho
      
      prob0 = (X + x)/(N + n); prob0[prob0==0] = pseudo; prob0[prob0==1] = 1 - pseudo
      prob1s = x/(n + pseudo); prob1s[prob1s==0] = pseudo; prob1s[prob1s==1] = 1 - pseudo
      prob1c = X/(N + pseudo); prob1c[prob1c==0] = pseudo; prob1c[prob1c==1] = 1 - pseudo
      
      prob1s = pmax(prob1s,prob1c) # Min error rate is that of the population (one-sided test)
      nu0 = prob0 * rdisp; nu1s = prob1s * rdisp; nu1c = prob1c * rdisp; 
      
      # Likelihood-Ratio Tests
      LL = logbb(x, n, nu0, rdisp) + logbb(X, N, nu0, rdisp) - logbb(x, n, nu1s, rdisp) - logbb(X, N, nu1c, rdisp)
      pvals = pchisq(-2*LL, df=1, lower.tail=F) / 2 # We divide by 2 as we are performing a 1-sided test
      # Saving the result
      pval_mat[k,s] = pvals
    } 
  }
  pval_mat = pval_mat[((rowSums(norm_NV) / rowSums(norm_NR)) < 0.2) & (NV[, matched_normal]/NR[, matched_normal] < 0.2),] #remove variants seen at a VAF of >0.2 in the normal reference panel or patient's matched normal, these end up being artefactual, even if they are seen at a significantly higher VAF in the tumour samples
  if(is.null(save)){
    return(pval_mat)
  }else{
    write.table(pval_mat, save)
  }
}

#-------------------------------------------------
# Input
#-------------------------------------------------

#setwd('/lustre/scratch117/casm/team294/to3/testes/tumour/shearwater_like_indels/20200523')
#options(stringsAsFactors = F)

#create list of patients
patient.list <- read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/cgpVAF/SNVs/patients_all.txt', header = F)[,1]

# Vector of normal reference samples, no GCT
normal_samples = read.table("/lustre/scratch117/casm/team294/to3/testes/tumour/shearwater_like_indels/20200513/rand_norm_reference.txt")[,1] 

# Vector of all tumour samples, used to create a per patient sample list
tumour_samples = read.table("/lustre/scratch117/casm/team294/to3/testes/tumour/shearwater_like_indels/20200513/gct_samples.txt")[,1]

for (i in unique(substr(tumour_samples, 0, 7))){
  patient.samples = tumour_samples[grepl(paste0(i), tumour_samples)]
  patient.samples = data.frame(patient.samples)
  write.table(patient.samples, paste0(i, '/samples.txt'), sep = '\t', col.names = F, row.names = F, quote = F)
}

#generate a list of mutations per patient in the format Chr_Ref_Pos_Alt
patient_muts_input = unlist(list.files('/lustre/scratch117/casm/team294/to3/testes/tumour/cgpVAF_phase2/', pattern = paste0('_indel_vaf_nohash.tsv')))

for (i in unique(substr(tumour_samples, 0, 7))){
  data <- read.table(paste0(i, '/', i, '.indels.bed'), header = F, sep = '\t')
  Muts = paste(data$V1,data$V2,data$V3,data$V4,sep="_")
  write.table(Muts, paste0(i, '/All_mutations_filtered.txt'), sep = '\t', col.names = F, row.names = F, quote = F)
}

# read in the list of study patients
patient.list <- read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/cgpVAF/SNVs/patients_all.txt', header = F)[,1]

#-------------------------------------------------
# Run
#-------------------------------------------------

for (patient in patient.list){
  shearwater_indel_probability(patient=patient,save=paste0(patient, "/shearwater_pval_indel_mat.txt"), rho = NULL) 
  print(patient)
}

#get an idea of the mutation burden per sample after this filtering step
post_swater_mb <- c()
for (patient in patient.list){
  pval_mat = read.table(paste0(patient, "/shearwater_pval_indel_mat.txt"), sep = ' ', row.names = 1, header = T)
  qval_mat = apply(pval_mat, 2, function(x) p.adjust(x, method="BH", n = length(as.matrix(pval_mat))))
  post_swater_mb <- c(post_swater_mb, colSums(qval_mat <= 0.001, na.rm = T))
}

#BH correction for multiple testing
for (patient in patient.list){
  pval_mat = read.table(paste0(patient, "/shearwater_pval_indel_mat.txt"), row.names = 1, sep = ' ', header = T)
  qval_mat = apply(pval_mat,2,function(x) p.adjust(x,method="BH",n = length(as.matrix(pval_mat))))
  write.table(qval_mat, paste0(patient, "/shearwater_qval_indel_mat.txt"), row.names = T, sep = '\t', quote = F)
}

#filter vcf to leave only variants that pass the site specific error threshold

library(VariantAnnotation)

for (patient in patient.list){
  qval_mat = read.table(paste0(patient, "/shearwater_qval_indel_mat.txt"), row.names = 1, sep = '\t', header = T)
  #qval_mat[is.na(qval_mat)] = 0.5 #replaces NA values should no longer be required
  samples = read.table(paste0(patient, "/samples.txt"), sep = '\t', header = F)[,1]
  for (sample in samples){
    data = readVcf(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/file_repo/SSMs/matched/', sample, '_indels_geno.final.vcf'))
    reads = cbind((geno(data)$MTR)[,2], (geno(data)$WTR)[,2])
    Var = data.frame(Chr = as.character(seqnames(rowRanges(data))), Pos=start(ranges(rowRanges(data))),
                     Ref=as.character(ref(data)))
    Alt_tmp = CharacterList(alt(data))
    Var$Alt = as.character(unlist(Alt_tmp))
    Var$NR = rowSums(reads)
    Var$NV = reads[,1]
    Var$VAF=Var$NV/Var$NR
    Var$ID = paste(Var$Chr, Var$Pos, Var$Ref, Var$Alt, sep = '_')
    qval_patient_sig_muts = row.names(qval_mat[qval_mat[,sample] <= 0.001,])
    swater_filtered_muts = Var[Var$ID %in% qval_patient_sig_muts,]
    write.table(swater_filtered_muts, paste0(patient, '/', sample, '_post_swater_final_indels.txt'), sep = '\t', quote = F, row.names = F)
  }
}

```

The remaining filtering steps for indels are described in the Methods section.

#Mutational signature extraction

Two orthogonal approaches were chosen - non-negative matrix factorization (Sigprofiler v1.015) and the hierarchical Dirchlet process (HDP v0.1.5). SigProfilerExtractor was run with a minimum and maximum of 2 and 20 signatures respectively. All other parameters were as per the default settings. The steps taken whilst running HDP are outlines below.

##HDP_0.1.5 with priors

This code is based off of scripts previously written by Tim Coorens and Daniel Leongamornlert.

###Parts A, B & C
PART A : Arrange the prior sigs
PART B : Set up HDP
PART C : Run posterior sampling chains

```{r}

#/nfs/users/nfs_t/to3/scripts/hdp_prior_single_chain.R

options(stringsAsFactors = F)
library(hdp)
lower_threshold=100

n=as.numeric(commandArgs(T)[1])
mutations=read.table("trinuc_mut_mat.txt")
key_table=read.table("key_table.txt")

ref = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/reformatted_cosmic_sig_v3.txt', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
ref = t(ref)
cosmic.sigs = read.csv('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/PCAWG_sigProfiler_SBS_signatures.csv', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
row.names(cosmic.sigs) = row.names(ref)
prior_sigs = as.matrix(cosmic.sigs)

# number of prior signatures to condition on (65)
nps <- ncol(prior_sigs)

# take a look
#prior_sigs[1:6,1:6]
# columns should sum to one
colSums(prior_sigs)

# have a look at the first two prior sigs:
#barplot(prior_sigs[,1], las=2, cex.names=0.5)
#barplot(prior_sigs[,2], las=2, cex.names=0.5)

#If requiring a minimum number of mutations:
sample_remove=rownames(mutations)[rowSums(mutations)<lower_threshold]
mutations=mutations[!rownames(mutations)%in%sample_remove,]
key_table=key_table[!key_table$Sample%in%sample_remove,]

#Hierarchy is set per patient, can change if wanted
freq = table(key_table$Patient)

nps <- ncol(prior_sigs)

luad_prior <- hdp_prior_init(prior_distn = prior_sigs, # matrix of prior sigs
                             prior_pseudoc = rep(1000, nps), # pseudocount weights
                             hh=rep(1, 96), # uniform prior over 96 categories
                             alphaa=c(1, 1), # shape hyperparams for 2 CPs
                             alphab=c(1, 1)) # rate hyperparams for 2 CPs
#luad_prior
#numdp(luad_prior)
#pseudoDP(luad_prior)
#conparam(luad_prior)
#ppindex(luad_prior)
#cpindex(luad_prior)
#dpstate(luad_prior) # 2 for active node, 1 for frozen, 0 for heldout

# make two more CPs available for the data we will add
luad_prior <- hdp_addconparam(luad_prior,
                              alphaa = rep(1,length(freq)+2), # shape hyperparams for x new CPs
                              alphab = rep(1,length(freq)+2)) # rate hyperparams for x new CPs

luad_prior <- hdp_adddp(luad_prior,
                        numdp = nrow(mutations) + 1,
                        ppindex = c(1, rep(1+nps+1:(length(freq)), times=freq)),
                        cpindex = c(3, rep(4:(length(freq)+3), times=freq)))

# assign the data to the relevant DP nodes
luad_prior <- hdp_setdata(luad_prior,
                          dpindex = (1+nps+1)+1:nrow(mutations), 
                          mutations) # mutation counts in all GCTs


hdp_activated <- dp_activate(luad_prior, 
                             dpindex = (1+nps+1)+0:nrow(mutations), 
                             initcc = nps+5,
                             seed = n * 1000)

chain=hdp_posterior(hdp_activated,
                    burnin=40000,
                    n = 200,
                    seed = n * 1000,
                    space = 300,
                    cpiter = 3)

saveRDS(chain,paste0("hdp_prior_chain_",n,".Rdata"))

```

```{bash}
#in directory with relevant files

export PATH=/software/R-3.6.1/bin:${PATH}
export R_HOME=$(R RHOME)
export R_LIBS_USER="~/custom_libraries/R/x86_64-pc-linux-gnu-library/3.6_hdp"

for n in $(seq 1 20);
do
bsub -o $PWD/log.%J -e $PWD/err.%J -q normal -R'select[mem>10000] rusage[mem=10000]' -M10000 -J $n /software/R-3.6.1/bin/Rscript /nfs/users/nfs_t/to3/scripts/hdp_prior_single_chain.R $n
done 

```

###Parts D & E
PART D : Review diagnostic plots
PART E : Extract consensus components / signatures

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/signature_extraction/hdp_0.1.5')

options(stringsAsFactors = F)
library(hdp)

chlist <- vector("list", 20)
for (i in 1:20){
  if(file.exists(paste0("hdp_prior_chain_",i,".Rdata"))){
    chlist[[i]] <- readRDS(paste0("hdp_prior_chain_",i,".Rdata"))
  }
}
if(any(unlist(lapply(chlist,is.null)))) chlist=chlist[-which(unlist(lapply(chlist,is.null)))]

luad_multi <- hdp_multi_chain(chlist)
pdf("QC_plots_chain.pdf") 
par(mfrow=c(2,2), mar=c(4, 4, 2, 1))
p1 <- lapply(chains(luad_multi), plot_lik, bty="L", start=1000)
p2 <- lapply(chains(luad_multi), plot_numcluster, bty="L")
p3 <- lapply(chains(luad_multi), plot_data_assigned, bty="L")
dev.off()

#need to remove chains 4, 5, 6, 9, 12, 16, 17

chlist <- vector("list", 20)
for (i in c(1:3, 7, 8, 10, 11, 13:15, 18:20)){
  if(file.exists(paste0("hdp_prior_chain_",i,".Rdata"))){
    chlist[[i]] <- readRDS(paste0("hdp_prior_chain_",i,".Rdata"))
  }
}
if(any(unlist(lapply(chlist,is.null)))) chlist=chlist[-which(unlist(lapply(chlist,is.null)))]

luad_multi <- hdp_multi_chain(chlist)

pdf("QC_plots_chain_after_review.pdf") 
par(mfrow=c(2,2), mar=c(4, 4, 2, 1))
p1 <- lapply(chains(luad_multi), plot_lik, bty="L", start=1000)
p2 <- lapply(chains(luad_multi), plot_numcluster, bty="L")
p3 <- lapply(chains(luad_multi), plot_data_assigned, bty="L")
dev.off()

luad_multi <- hdp_extract_components(luad_multi)
saveRDS(luad_multi,"HDP_multi_chain.Rdata")

#luad_multi = readRDS('HDP_multi_chain.Rdata')

numcomp(luad_multi) #6
prop.ex(luad_multi) #0.91

pdf("muts_attributed.pdf")
plot_comp_size(luad_multi, bty="L")
dev.off()

# plot components / signatures
# pick your colours
mut_colours=c("dodgerblue","black","red","grey70","olivedrab3","plum2")
# labels along bottom x-axis
trinuc_context <- sapply(strsplit(colnames(mut_count), '\\.'), `[`, 4)
# group labels along the top (and controls colour grouping)
group_factor <- as.factor(rep(c("C>A", "C>G", "C>T", "T>A", "T>C", "T>G"),
                              each=16))

ref = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/reformatted_cosmic_sig_v3.txt', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
ref = t(ref)
cosmic.sigs = read.csv('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/PCAWG_sigProfiler_SBS_signatures.csv', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
row.names(cosmic.sigs) = row.names(ref)
prior_sigs = as.matrix(cosmic.sigs)

# number of prior signatures to condition on (65)
nps <- ncol(prior_sigs)

pdf(paste0("hdp_components_with_priors.pdf"),width=12,height=4)
plot_comp_distn(luad_multi, cat_names=trinuc_context,
                grouping=group_factor, col=mut_colours,
                col_nonsig="grey97", show_group_labels=TRUE)
dev.off()

lower_threshold = 100
mutations=read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/HDP/input/trinuc_mut_mat.txt")
key_table=read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/HDP/input/key_table.txt")
sample_remove=rownames(mutations)[rowSums(mutations)<lower_threshold]
mutations=mutations[!rownames(mutations)%in%sample_remove,]
key_table=key_table[!key_table$Sample%in%sample_remove,]


pdf("signature_attribution.pdf",width=10,height=8)
plot_dp_comp_exposure(luad_multi, dpindices = (1+nps+1)+1:nrow(mutations), incl_nonsig = T,
                      col=c('black', RColorBrewer::brewer.pal(numcomp(luad_multi), "Set1")), cex.names=0.8,
                      ylab_exp = 'Signature exposure', leg.title = 'Signature')
dev.off()

mean_assignment=as.data.frame(comp_dp_distn(luad_multi)$mean)
write.table(mean_assignment,"mean_assignment_prior_hdp.txt")
mean_sigs=as.data.frame(t(comp_categ_distn(luad_multi)$mean))
write.table(mean_sigs,"hdp_prior_sigs.txt")

```

###Deconvolute 'novel' signatures

```{r}

#Deconvolute HDP signatures into reference signatures and arrive at final set

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/signature_extraction/hdp_0.1.5')

options(stringsAsFactors = F)
library(hdp)
library(RColorBrewer)
library(lsa)
library(lattice)

mut.cols = rep(c("dodgerblue","black","red","grey70","olivedrab3","plum2"),each=16)

#Load HDP signatures
mean_sigs = read.table("hdp_prior_sigs.txt", row.names = 1, header = T)
hdp_sigs = mean_sigs
#Load reference signatures, need formatting
ref = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/reformatted_cosmic_sig_v3.txt', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
ref = t(ref)
cosmic.sigs = read.csv('/lustre/scratch119/casm/team294rr/to3/testes/tumour/MutationalPatterns/PCAWG_sigProfiler_SBS_signatures.csv', header = T, stringsAsFactors = F, row.names = 1, check.names = F)
row.names(cosmic.sigs) = row.names(ref)
prior_sigs = as.matrix(cosmic.sigs)
ref = prior_sigs

#Assess cosine similarities for all unassigned signatures (X0, N1, N2)
cosine_matrix=data.frame(matrix(nrow=ncol(hdp_sigs[, c('X0', 'N1', 'N2')]), ncol=ncol(ref)))
rownames(cosine_matrix)=colnames(hdp_sigs[, c('X0', 'N1', 'N2')])
colnames(cosine_matrix)=colnames(ref)

for (n in 1:nrow(cosine_matrix)) {
  for (m in 1:ncol(cosine_matrix)) {
    cosine_matrix[n,m] <- cosine(x=hdp_sigs[,rownames(cosine_matrix)[n]],
                                 y=ref[,colnames(cosine_matrix)[m]])
  }
}

write.table(cosine_matrix, "Cosine_similarities_unassigned_to_priors.txt",sep="\t",quote=F)

#cosine_matrix = read.table('Cosine_similarities_unassigned_to_priors.txt', header = T, sep = '\t')
#rownames(cosine_matrix) = c('X0', 'N1', 'N2')
hdp_sigs = hdp_sigs[, c('X0', 'N1', 'N2')]

#plot output
pdf("cosine_similarities_unassigned_to_priors.pdf", height=5, width=15)
color.palette = colorRampPalette(c("white", "orange", "purple"))
levelplot(t(cosine_matrix[dim(cosine_matrix)[1]:1,]),col.regions=color.palette, aspect="fill", scales=list(x=list(rot=90)))
dev.off()

#select which signatures to decompose into reference signatures
sigs_to_decompose=rowSums(cosine_matrix>0.85)
for(n in 1:nrow(cosine_matrix)){
  print(paste0(rownames(cosine_matrix)[n],": ",paste(colnames(cosine_matrix)[cosine_matrix[n,]>0.7],collapse=",")))
}

#N1 has large error bars and does not look stable across the chains

#"0: SBS3,SBS5,SBS40" stable, flat signature, closest to SBS5, followed by SBS40, then SBS3
#"N1: SBS3,SBS5,SBS39,SBS40" assortment of flat signature, unstable
#"N2: SBS31,SBS35" platinum signature

#First iteration; decomposed hdp sigs into all suspected sigs 
gdsigs=c("SBS3", "SBS5", "SBS31", "SBS35", "SBS39", "SBS40") #0.7 threshold

signatures=t(ref[,gdsigs])
sample_list= c('X0', 'N1', 'N2')
profiles=hdp_sigs[,sample_list]

signature_fraction = matrix(NA,nrow=nrow(signatures),ncol=length(sample_list))
rownames(signature_fraction) = rownames(signatures)
colnames(signature_fraction) = sample_list
maxiter <- 1000

for (j in 1:length(sample_list)) {
  freqs = profiles[,j]
  freqs[is.na(freqs)] = 0
  # EM algorithm with to estimate the signature contribution
  alpha = runif(nrow(signatures)); alpha=alpha/sum(alpha) # Random start (seems to give ~identical results)
  for (iter in 1:maxiter) {
    contr = t(array(alpha,dim=c(nrow(signatures),96))) * t(signatures)
    probs = contr/array(rowSums(contr),dim=dim(contr))
    probs = probs * freqs
    old_alpha = alpha
    alpha = colSums(probs)/sum(probs)
    if (sum(abs(alpha-old_alpha))<1e-5) {
      break
    }
  }
  # Saving the signature contributions for the sample
  print(j/length(sample_list))
  signature_fraction[,j] = alpha
}

#Plot initial deconvolution and save results
pdf("Deconvolution_hdp_sigs_after_prior_R1.pdf", height=5, width=10)
color.palette = colorRampPalette(c("white", "orange", "purple"))
levelplot((signature_fraction[nrow(signature_fraction):1,]),col.regions=color.palette, aspect="fill", scales=list(x=list(rot=90)))
dev.off()

write.table(signature_fraction, "hdp_unassigned_sigs_broken_down_into_cosmic_sigs.txt", sep="\t", col.names=T, row.names = T, quote=F)

```

#Aggregated copy number plotting - study cohort vs TCGA

Turn our battenberg files into a long segment format

```{bash}

cd lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/CNA/

for f in $(ls *.battenberg.subclones.txt);
do
filename=$(basename -- "$f")
sampleID="${filename%.battenberg.subclones.txt}"
python /nfs/users/nfs_t/to3/scripts/copynumber.py -i $sampleID.battenberg.subclones.txt -c $sampleID.battenberg.rho_and_psi_updated.txt -o . -r
done

```


- In-house data binned into 10kb chunks and the copy number across that chunk used according to the Battenberg subclones file.
- The median CN for each bin per tumour is taken across all invasive samples with a purity >40%.
- The average of this composite per tumour CN is then taken as the snapshot for the cohort as a whole.
- Autosomal chromosomes only as we have some ovarian tumours too.

```{bash}

cd /lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp

bsub -G team294-vwork -o $PWD/log.%J -e $PWD/err.%J -q yesterday -R'select[mem>30000] rusage[mem=30000]' -M30000 -J 'gct_cn' /software/R-3.6.1/bin/Rscript lcm_gct_WG_CN.R

```


```{r}

#lcm_gct_WG_CN.R
setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/')

library(plyr)

#locate copy number files and the eligible samples
project_dir <- "/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/CNA/"
dna_cuts = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_DNA_supplementary_data.txt', header = T, sep = '\t', stringsAsFactors = F)
pure_dna_cuts = dna_cuts[dna_cuts$purity > 0.4 & !is.na(dna_cuts$purity) & dna_cuts$Histology != 'GCNIS',] #pure, invasive only
samples <- pure_dna_cuts$Sample #all samples from all donors
donors <- unique(substr(samples,1,7)) #all donors

#set bin size
bin_size <- 10000
filter_size = 10000 #remove CN changes smaller than this
adjust = bin_size - 1 #for arranging segments

#chromosome lengths, each will be rounded up to the nearest multiple of 10,000
chrom_sizes <- read.table('/lustre/scratch119/casm/team294rr/rs30/Testicular_project/Final_Analyses/metadata/male.hg19.chrom.sizes.tsv', header = F, sep = '\t', stringsAsFactors = F)
chrom_sizes[,2] = round_any(chrom_sizes[,2], bin_size, ceiling) #chunk each chromosome into 10kb bins in order to accurate characterise the CN at the beginning and ends of chromosomes, has to be ceiling to capture the ends
chrom_sizes[,3] <- cumsum(as.numeric(chrom_sizes[,2]))
chrom_sizes = chrom_sizes[!chrom_sizes$V1 %in% c('chrM', 'chrX', 'chrY'),]
chrom_sizes$V4 = c(0, chrom_sizes$V3[1:(length(chrom_sizes$V3) - 1)])
chrom_sizes$V1 = unlist(lapply(strsplit(chrom_sizes$V1, 'chr'), '[[', 2))
names(chrom_sizes) = c('chr', 'chr_length', 'cumsum_length', 'prior_wg_length')

#create array with per donor and per 10kb bin cells
cn_arr <- array(data = NA, dim = c((chrom_sizes[22, 'cumsum_length'] / bin_size), length(donors)))
colnames(cn_arr) = donors
cn_df = data.frame(cn_arr)

for(i in 1:length(donors)) {
  print(donors[i])
  donor_samples = samples[grepl(donors[i], samples)]
  donor_sample_arr = array(data = NA, dim = c((chrom_sizes[22, 'cumsum_length'] / bin_size), length(donor_samples))) #need copy number per site per sample first before averaging across donor and ultimately across all tumours
  colnames(donor_sample_arr) = donor_samples
  donor_sample_df = data.frame(donor_sample_arr)
  for(j in 1:length(donor_samples)){
    if(file.exists(paste0(project_dir,"/",donor_samples[j],"_segments.txt"))){
    print(donor_samples[j])
    segments <- read.table(paste0(project_dir,"/",donor_samples[j],"_segments.txt"), sep = "\t", stringsAsFactors = F, header = T)
    segments = segments[segments$chromosome %in% c(1:22),] #avoid sex chromosomes due to sampling of men and women
    segments = segments[order(segments$clonal_frequency, decreasing = T),] #order by cn state with greatest clonal frequency first
    segments = segments[!duplicated(segments[,1:3]),] #only keep the integer value for the copy number state with the highest clonal frequency
    segments = segments[, c('chromosome', 'start', 'end', 'copy_number')]
    names(segments) = c('chr', 'startpos', 'endpos', 'ntot')
    
  
    segments = segments[(segments$endpos - segments$startpos) > filter_size,] #minimum copy number segment size
    segments$startpos = round_any(segments$startpos, bin_size, round) #round start and end positions to nearest 10kb
    segments$endpos = round_any(segments$endpos, bin_size, round)
    segments$startcumpos <- segments$startpos #treating genome as one continuous length, generate cumsum postions
    segments$endcumpos <- segments$endpos
   
    
   for(k in 1:nrow(segments)) { #annotate segments as cumulative positions
        segments$startcumpos[k] = segments$startpos[k] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == segments$chr[k])]
        segments$endcumpos[k] <- segments$endpos[k] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == segments$chr[k])]
    }
    
    sample_vec = c()
    for(m in 1:nrow(donor_sample_df)){ #get copy number estimates per 10000bp bin
      if(nrow(segments[segments$startcumpos <= (bin_size * (m - 1)) & segments$endcumpos >= (bin_size * m),]) > 0) sample_vec = c(sample_vec, segments[segments$startcumpos <= (bin_size * (m - 1)) & segments$endcumpos >= (bin_size * m),]$ntot)
      else {
        sample_vec = c(sample_vec, NA)
      }
    }
    donor_sample_df[, donor_samples[j]] = sample_vec
    }
  }
  cn_df[, donors[i]] = apply(donor_sample_df, 1, median, na.rm = T) #median per tumour for that bin
}
  
write.table(cn_df, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/lcm_gct_cn_10000_bp_bin_per_GCT_WG_median.txt', col.names = T, row.names = F, sep = '\t', quote = F) 

```

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/')

cn_df = read.table('lcm_gct_cn_10000_bp_bin_per_GCT_WG_median.txt', header = T, sep = '\t', stringsAsFactors = F)
cn_df$start = c(1:nrow(cn_df))*10000 - 9999
cn_df$end = c(1:nrow(cn_df))*10000

#adjust for ploidy, to express copy number relative to the overall ploidy 
dna_cuts = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_DNA_supplementary_data.txt', header = T, sep = '\t', stringsAsFactors = F)
pure_dna_cuts = dna_cuts[dna_cuts$purity > 0.4 & !is.na(dna_cuts$purity) & dna_cuts$Histology != 'GCNIS',] #pure, invasive only
lcm_ploidy_table = aggregate(ploidy ~ Case_ID, pure_dna_cuts, median)
row.names(lcm_ploidy_table) = lcm_ploidy_table$Case_ID
cn_df[,1:12] = sweep(cn_df[,1:12], 2, lcm_ploidy_table[colnames(cn_df[,1:12]),]$ploidy, FUN = '-', check.margin = T) #adjust the copy number calls for each tumour by their median ploidy

#add further statistics
#cn_df$all_samples = apply(cn_df[,1:12], MARGIN = 1, FUN = median, na.rm = T) 
cn_df$all_samples = rowMeans(cn_df[,1:12], na.rm = T)
cn_df$postp_samples = rowMeans(cn_df[,c(1:3, 5:12)], na.rm = T) 
cn_df$uq_all_samples = apply(cn_df[,c(1:12)], MARGIN = 1, FUN = function(z) {quantile(z, 0.75, na.rm = T)})
cn_df$lq_all_samples = apply(cn_df[,c(1:12)], MARGIN = 1, FUN = function(z) {quantile(z, 0.25, na.rm = T)})
cn_df$uq_postp_samples = apply(cn_df[,c(1:3, 5:12)], MARGIN = 1, FUN = function(z) {quantile(z, 0.75, na.rm = T)})
cn_df$lq_postp_samples = apply(cn_df[,c(1:3, 5:12)], MARGIN = 1, FUN = function(z) {quantile(z, 0.25, na.rm = T)})

write.table(cn_df, 'lcm_gct_cn_10000_bp_bin_per_GCT_WG_median_extra_cols.txt', col.names = T, row.names = F, sep = '\t', quote = F)

```

- To derive the IQR from the TCGA data, we need to do something similar with the ASCAT segments, also from the hg19 build using the same purity cut-off
- Using latest ASCAT segmentation and ploidy estimates provided by Peter van Loo and his group for the TCGA data.
- Limited to tumours which were over 40% pure.

```{bash}

cd /lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp

bsub -G team294-vwork -o $PWD/log.%J -e $PWD/err.%J -q yesterday -R'select[mem>30000] rusage[mem=30000]' -M30000 -J 'tcga_cn' /software/R-3.6.1/bin/Rscript tcga_gct_WG_CN.R

```

```{r}

#tcga_gct_WG_CN.R
setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/')

library(plyr)

#get metadata, including ploidy values for all eligible tumours
tcga_ploidy = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct/summary.ascatTCGA.penalty70.hg19.tsv', header = T, sep = '\t', stringsAsFactors = F)
tcga_ploidy = tcga_ploidy[(tcga_ploidy$cancer_type == 'TGCT') & (tcga_ploidy$pass == TRUE) & (tcga_ploidy$rep == TRUE) & (tcga_ploidy$purity > 0.4), ] #103 samples

#get copy segments for all tumours in a single dataframe
tcga_segment_files = list.files(path = '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct', pattern = 'segments.txt')
paste0(tcga_ploidy$name, '.segments.txt') %in% tcga_segment_files #all present in segment files

tcga_ascat_segments = c()
for(i in tcga_ploidy$name){
  tumour_file = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct/', i, '.segments.txt'), header = T, sep = '\t', stringsAsFactors = F)
  tcga_ascat_segments = rbind(tcga_ascat_segments, tumour_file)
}

names(tcga_ascat_segments) = c('tumour', 'chr', 'start', 'end', 'majA', 'minA')
tcga_ascat_segments$ntot = tcga_ascat_segments$majA + tcga_ascat_segments$minA

samples <- unique(tcga_ascat_segments$tumour)

#set bin size
bin_size <- 10000
filter_size = 10000 #remove CN changes smaller than this
adjust = bin_size - 1 #for arranging segments

#chromosome lengths, each will be rounded up to the nearest multiple of 10,000
chrom_sizes <- read.table('/lustre/scratch119/casm/team294rr/rs30/Testicular_project/Final_Analyses/metadata/male.hg19.chrom.sizes.tsv', header = F, sep = '\t', stringsAsFactors = F)
chrom_sizes[,2] = round_any(chrom_sizes[,2], bin_size, ceiling) #chunk each chromosome into 10kb bins in order to accurate characterise the CN at the beginning and ends of chromosomes, has to be ceiling to capture the ends
chrom_sizes[,3] <- cumsum(as.numeric(chrom_sizes[,2]))
chrom_sizes = chrom_sizes[!chrom_sizes$V1 %in% c('chrM', 'chrX', 'chrY'),]
chrom_sizes$V4 = c(0, chrom_sizes$V3[1:(length(chrom_sizes$V3) - 1)])
chrom_sizes$V1 = unlist(lapply(strsplit(chrom_sizes$V1, 'chr'), '[[', 2))
names(chrom_sizes) = c('chr', 'chr_length', 'cumsum_length', 'prior_wg_length')

tcga_ascat_segments = tcga_ascat_segments[(tcga_ascat_segments$end - tcga_ascat_segments$start) > filter_size,]
tcga_ascat_segments$mod_start = round_any(tcga_ascat_segments$start, bin_size, round) #fit genome to bins of 10,000bp
tcga_ascat_segments$mod_end = round_any(tcga_ascat_segments$end, bin_size, round)
tcga_ascat_segments = tcga_ascat_segments[tcga_ascat_segments$chr %in% c(1:22),]

tcga_ascat_segments$startcumpos = tcga_ascat_segments$mod_start
tcga_ascat_segments$endcumpos = tcga_ascat_segments$mod_end

for(j in 1:nrow(tcga_ascat_segments)) { #annotate segments as cumulative positions
        tcga_ascat_segments$startcumpos[j] = tcga_ascat_segments$mod_start[j] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == tcga_ascat_segments$chr[j])]
        tcga_ascat_segments$endcumpos[j] <- tcga_ascat_segments$mod_end[j] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == tcga_ascat_segments$chr[j])]
}

#create array per sample and per 10kb bin cells
tcga_cn_arr <- array(data = NA, dim = c((chrom_sizes[22, 'cumsum_length'] / bin_size), length(samples)))
colnames(tcga_cn_arr) = samples
tcga_cn_df = data.frame(tcga_cn_arr)
colnames(tcga_cn_df) = gsub('\\.', '-', colnames(tcga_cn_df))

for(k in 1:length(samples)){
  print(samples[k])
  for(m in 1:nrow(tcga_cn_df)){ #get copy number estimates per 10000bp bin
      if(nrow(tcga_ascat_segments[tcga_ascat_segments$startcumpos <= (10000 * (m - 1)) & tcga_ascat_segments$endcumpos >= (10000 * m) & tcga_ascat_segments$tumour == samples[k],]) > 0){
        tcga_cn_df[m, samples[k]] = tcga_ascat_segments[tcga_ascat_segments$startcumpos <= (10000 * (m - 1)) & tcga_ascat_segments$endcumpos >= (10000 * m) & tcga_ascat_segments$tumour == samples[k],]$ntot
    }
  }
}

write.table(tcga_cn_df, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/tcga_cn_10000_bp_GCT_WG_median.txt', col.names = T, row.names = F, sep = '\t', quote = F) 

```


```{r}

tcga_cn_df = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/tcga_cn_10000_bp_GCT_WG_median.txt', header = T, sep = '\t', stringsAsFactors = F, check.names = F)

#read in ploidy values
#identify samples with RNA seq data
tcga_ploidy = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct/summary.ascatTCGA.penalty70.hg19.tsv', header = T, sep = '\t', stringsAsFactors = F)
tcga_ploidy = tcga_ploidy[(tcga_ploidy$cancer_type == 'TGCT') & (tcga_ploidy$pass == TRUE) & (tcga_ploidy$rep == TRUE) & (tcga_ploidy$purity > 0.4), ] #103 samples
row.names(tcga_ploidy) = tcga_ploidy$name

#substract the ploidy from the counts for each tumour
tcga_cn_df = sweep(tcga_cn_df, 2, tcga_ploidy[colnames(tcga_cn_df),]$ploidy, FUN = '-', check.margin = T)
#tcga_cn_df$all_samples = apply(tcga_cn_df[, 1:103], 1, median, na.rm = T)
tcga_cn_df$all_samples = rowMeans(tcga_cn_df[, 1:103], na.rm = T)
tcga_cn_df$start = c(1:nrow(tcga_cn_df))*10000 - 9999
tcga_cn_df$end = c(1:nrow(tcga_cn_df))*10000
tcga_cn_df$uq = apply(tcga_cn_df[,1:103], MARGIN = 1, FUN = function(z) {quantile(z, 0.75, na.rm = T)})
tcga_cn_df$lq = apply(tcga_cn_df[,1:103], MARGIN = 1, FUN = function(z) {quantile(z, 0.25, na.rm = T)})

write.table(tcga_cn_df, 'tcga_cn_10000_bp_GCT_WG_median_extra_cols.txt', col.names = T, row.names = F, sep = '\t', quote = F)

```

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/')

library(ggplot2)
library(ggpubr)

cn_df = read.table('lcm_gct_cn_10000_bp_bin_per_GCT_WG_median_extra_cols.txt', header = T, sep = '\t', stringsAsFactors = F, check.names = F)
tcga_cn_df = read.table('tcga_cn_10000_bp_GCT_WG_median_extra_cols.txt', header = T, sep = '\t', stringsAsFactors = F, check.names = F)

chrom_sizes <- read.table('/lustre/scratch119/casm/team294rr/rs30/Testicular_project/Final_Analyses/metadata/male.hg19.chrom.sizes.tsv', header = F, sep = '\t', stringsAsFactors = F)
chrom_sizes[,2] = round_any(chrom_sizes[,2], bin_size, ceiling) #chunk each chromosome into 10kb bins in order to accurate characterise the CN at the beginning and ends of chromosomes, has to be ceiling to capture the ends
chrom_sizes[,3] <- cumsum(as.numeric(chrom_sizes[,2]))
chrom_sizes = chrom_sizes[!chrom_sizes$V1 %in% c('chrM', 'chrX', 'chrY'),]
chrom_sizes$V4 = c(0, chrom_sizes$V3[1:(length(chrom_sizes$V3) - 1)])
chrom_sizes$V1 = unlist(lapply(strsplit(chrom_sizes$V1, 'chr'), '[[', 2))
names(chrom_sizes) = c('chr', 'chr_length', 'cumsum_length', 'prior_wg_length')

#get genes loci, examine overall hotspots
data = read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/featureCounts/lcm_rna_featurecount_data_preQC_20201013.txt", header = T, sep = '\t', stringsAsFactors = F, check.names = T)
row.names(data) = data$Geneid
data$Start = as.numeric(unlist(lapply(strsplit(data$Start, ';'), '[[', 1)))
data$End = as.numeric(unlist(lapply(strsplit(data$End, ';'), '[[', 1)))
data$startcumpos = data$Start
data$endcumpos = data$End
data = data[data$Chr %in% c(1:22),] #only autosomal chromosomes

for(k in 1:nrow(data)) { #annotate segments as cumulative positions
    if(data$Chr[k] != "1"){
        data$startcumpos[k] = data$Start[k] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == data$Chr[k])]
        data$endcumpos[k] = data$End[k] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == data$Chr[k])]
      }
}
data = data[, c('Geneid', 'Chr', 'Start', 'End', 'startcumpos', 'endcumpos')]
genes_to_plot = data[c('KRAS', 'KIT', 'CCND2'),]
genes_to_plot$plot_pos = NA
for(i in 1:nrow(genes_to_plot)){
  genes_to_plot$plot_pos[i] = cn_df[cn_df$start < genes_to_plot$startcumpos[i] & cn_df$end > genes_to_plot$startcumpos[i], ]$postp_samples
}

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/lcm_gct_v_tcga_cn_20210209.pdf', height = 3, width = 12, useDingbats = F)
ggplot() + 
  scale_x_continuous(expand = c(0,0), breaks = c(chrom_sizes$cumsum_length - (chrom_sizes$chr_length / 2)), labels = c(1:22), limits = c(0, 2881130000)) +
  geom_ribbon(tcga_cn_df, mapping = aes(x = start, ymin = lq, ymax = uq), alpha = 0.4, colour = 'transparent', fill = '#999999') +
  geom_point(cn_df, mapping = aes(x = start, y = postp_samples), size = 0.5) + 
  geom_vline(xintercept = chrom_sizes$prior_wg_length) + labs(y = 'copy number - ploidy', x = 'chr') + theme_pubr() + theme(panel.border = element_rect(fill = NA)) + coord_cartesian(expand = F) + geom_hline(yintercept = 0, col = '#999999', lty = 2) + ylim(-2, 8) + geom_text(genes_to_plot, mapping = aes(x = startcumpos, y = plot_pos + 0.5), label = c('KRAS', 'KIT', 'CCND2'))
dev.off()
                                                                                                                                               
```

#Telomere content modelling

There are many biological and technical factors that influence telomere length estimates. We first review the raw content values and their association with metadata parameters before we construct a linear mixed effects model to adjust for factors that are not of interest.

```{r}
library(lme4)
library(lmerTest)
library(ggplot2)
library(ggpubr)
library(ggbeeswarm)
library(remef)

telomerehunter_df = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/telomerehunter/gct_telomerehunter_matrix_20201110.txt', sep = '\t', header = T, stringsAsFactors = F) #read in telomerehunter content values for cohort
telomerehunter_df = telomerehunter_df[telomerehunter_df$patientID != 'PD46967',] #excluded as DNA samples were not included in the study from this patient due to the lack of a matched normal

#make dysgerminoma/seminoma into one category so model as site is already another variable that will discriminate these tissues between testis and ovary
telomerehunter_df[telomerehunter_df$histology %in% c('Dysgerminoma', 'Seminoma'),]$histology = 'Dysgerminoma/Seminoma'

#plot the variables to assess whether they may impact telomere length

##platform effect
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = histology, y = tel_content, col = platform)) #hiseq seems to have slightly higher telomere lengths
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = histology, y = tel_content, col = platform, shape = site))
ggplot(telomerehunter_df[telomerehunter_df$histology == 'normal_ST',]) + geom_point(mapping = aes(x = patientID, y = tel_content, col = platform))
ggplot(telomerehunter_df[telomerehunter_df$histology == 'normal_ST',]) + geom_point(mapping = aes(x = age, y = tel_content, col = platform))
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour' & telomerehunter_df$platform == 'NovaSeq',]) + geom_point(mapping = aes(x = histology, y = tel_content, col = site)) #even in Novaseq only, the telomeres of ovarian tumours were shorter

##age
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = age, y = tel_content, col = histology)) #age effect not appreciable
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'normal',]) + geom_point(mapping = aes(x = age, y = tel_content, col = histology))

##coverage
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = coverage, y = tel_content, col = histology)) #coverage effect not overtly appreciable
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'normal',]) + geom_point(mapping = aes(x = coverage, y = tel_content, col = histology))

##diagnosis
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = histology, y = tel_content, col = diagnosis, shape = site)) #differences perhaps confounded by sequencing platform

#patient
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = histology, y = tel_content, col = patientID, shape = platform)) #reasonable consistency in telomere lengths across histologies between patients

##site
ggplot(telomerehunter_df[telomerehunter_df$normal.tumour == 'tumour',]) + geom_point(mapping = aes(x = site, y = tel_content, col = patientID, shape = platform)) #all ovarian telomeres are considerably shorter

#create basic models to compare
##null model
null_model = lmer(formula = tel_content ~ 0 + (1 | patientID), data = telomerehunter_df, REML = F)

##full model
full_model = lmer(formula = tel_content ~ 0 + (1 | patientID) + coverage + age + platform + diagnosis + histology + site, data = telomerehunter_df, REML = F)

##compare null and full
anova(null_model, full_model) #full better than null, fixed effects do seem to impact telomere length, as expect

##optimise model using stepwise algorithm
step(full_model) #optimum model = tel_content ~ (1 | patientID) + platform + diagnosis + histology + site - 1, age and coverage not very informative
final_model = lmer(formula = tel_content ~ (1 | patientID) + platform + diagnosis + histology + site - 1, data = telomerehunter_df, REML = F)
anova(null_model, full_model, final_model) 
summary(final_model)

#partial regression - to remove the effect of patient and sequencing platform
full_model.no_patientID_platform_partial = remef(final_model, fix = c("platformHiSeq", "platformNovaSeq"), ran = 'all')
telomerehunter_df$tel_content_partial = full_model.no_patientID_platform_partial
telomerehunter_df$diagnosis_plot = telomerehunter_df$diagnosis
telomerehunter_df[telomerehunter_df$diagnosis %in% c('blood', 'normal_ST'),]$diagnosis_plot = 'Normal'

telomerehunter_df$histology = factor(telomerehunter_df$histology, levels = c('blood', 'normal_ST', 'GCNIS', 'Dysgerminoma/Seminoma', 'Embryonal_carcinoma', 'Yolk_sac_tumour', 'Neuroectoderm', 'Cartilage_teratoma', 'Glandular_teratoma', 'Malignant_stroma', 'Smooth_muscle_teratoma', 'Syncytiotrophoblasts'))
levels(telomerehunter_df$histology) = c('Blood', 'Seminiferous tubule', 'GCNIS', 'Dysgerminoma/Seminoma', 'Embryonal carcinoma', 'Yolk sac tumour', 'Neuroepithelium', 'Cartilage teratoma', 'Glandular teratoma', 'Malignant stroma', 'Smooth muscle teratoma', 'Syncytiotrophoblasts')

telomerehunter_df$diagnosis_plot_reannot = telomerehunter_df$diagnosis_plot
telomerehunter_df[telomerehunter_df$diagnosis_plot == 'Seminoma' & telomerehunter_df$histology == 'Dysgerminoma/Seminoma',]$diagnosis_plot_reannot = 'Seminoma, pure seminoma'
telomerehunter_df[telomerehunter_df$diagnosis_plot == 'Post-pubertal mixed germ cell tumour' & telomerehunter_df$histology == 'Dysgerminoma/Seminoma',]$diagnosis_plot_reannot = 'Dysgerminoma/Seminoma, NSGCT'
telomerehunter_df[telomerehunter_df$diagnosis_plot == 'Post-pubertal mixed germ cell tumour' & !(telomerehunter_df$histology %in% c('GCNIS', 'Dysgerminoma/Seminoma')),]$diagnosis_plot_reannot = 'Other tissues, NSGCT'
telomerehunter_df[telomerehunter_df$diagnosis_plot == 'Seminoma' & telomerehunter_df$histology == 'GCNIS',]$diagnosis_plot_reannot = 'GCNIS, pure seminoma'
telomerehunter_df[telomerehunter_df$diagnosis_plot == 'Post-pubertal mixed germ cell tumour' & telomerehunter_df$histology == 'GCNIS',]$diagnosis_plot_reannot = 'GCNIS, NSGCT'

telomerehunter_df$diagnosis_plot_reannot = factor(telomerehunter_df$diagnosis_plot_reannot, levels = c('Normal', 'GCNIS, pure seminoma', 'GCNIS, NSGCT', 'Seminoma, pure seminoma', 'Dysgerminoma/Seminoma, NSGCT', 'Pre-pubertal yolk sac tumour', 'Other tissues, NSGCT'))
levels(telomerehunter_df$diagnosis_plot_reannot) = c('Normal', 'GCNIS, pure seminoma', 'GCNIS, NSGCT', 'Seminoma, pure seminoma', 'Dysgerminoma/Seminoma, NSGCT', 'Prepubertal YST', 'Other tissues, NSGCT')
telomerehunter_df$site = as.factor(telomerehunter_df$site)
levels(telomerehunter_df$site) = c('Blood', 'Ovary', 'Testis')

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_telomerecontent_by_tissue_by_tumour.pdf', width = 8, height = 6, useDingbats = F)
ggplot(telomerehunter_df) + 
  geom_quasirandom(mapping = aes(x = diagnosis_plot_reannot, y = tel_content_partial, col = diagnosis_plot_reannot), cex = 2) + 
  geom_pointrange(mapping = aes(x = diagnosis_plot_reannot, y = tel_content_partial),
                  stat = "summary", shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill="black", cex = 1) +
  facet_grid(. ~ site, scales = 'free_x', space = 'free_x') + theme_pubr(legend = 'right') + 
  labs(y = 'Normalised telomere content', x = '') + theme(axis.text.x = element_text(angle = 315, vjust = 1, hjust = 0), strip.background = element_blank(), strip.text.x = element_text(size = 12)) + scale_color_manual(values = c('#999999', as.vector(get_palette(palette = "Dark2", 6))), name = 'Category')
dev.off()

#get stats to place on plot
wilcox.test(telomerehunter_df[telomerehunter_df$diagnosis_plot_reannot == 'Dysgerminoma/Seminoma, NSGCT' & telomerehunter_df$site == 'Ovary', ]$tel_content_partial, telomerehunter_df[telomerehunter_df$diagnosis_plot_reannot == 'Dysgerminoma/Seminoma, NSGCT' & telomerehunter_df$site == 'Testis', ]$tel_content_partial, alternative = 'two.sided') #W = 0, p-value = 0.009524 **
wilcox.test(telomerehunter_df[telomerehunter_df$diagnosis_plot_reannot == 'Other tissues, NSGCT' & telomerehunter_df$site == 'Ovary', ]$tel_content_partial, telomerehunter_df[telomerehunter_df$diagnosis_plot_reannot == 'Other tissues, NSGCT' & telomerehunter_df$site == 'Testis', ]$tel_content_partial, alternative = 'two.sided') #W = 0, p-value = 3.052e-08 ***

#get number of genomes that support each category
aggregate(tel_content_partial ~ diagnosis_plot_reannot + site, telomerehunter_df, length)

```

#Calculate multiplicity of driver substitutions

Used to time their occurence relative to WGD

```{r}

manifest = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_DNA_supplementary_data.txt', header = T, sep = '\t', stringsAsFactors = F)
row.names(manifest) = manifest$Sample

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/SNV')

#12 25398284 #both KRAS mutations
#14	105243048 #AKT1 mutation

#"PD45543a_lo0002" "PD45543a_lo0004" "PD45544a_lo0001" "PD45544a_lo0003" "PD45544a_lo0006" "PD45544a_lo0010" "PD45544a_lo0011" "PD45545a_lo0004" "PD45545a_lo0007"

driver_muts = c()
for(i in manifest[manifest$Case_ID %in% c('PD45543', 'PD45544', 'PD45545') & manifest$Battenberg == 1,]$Sample){
  mt_df = read.table(paste0(i, '_post_swater_final_snvs.txt'), header = T, sep = '\t', stringsAsFactors = F)
  mt_df = mt_df[(mt_df$Chr == 12 & mt_df$Pos == 25398284) | (mt_df$Chr == 14 & mt_df$Pos == 105243048),]
  mt_df$sample = i
  driver_muts = rbind(driver_muts, mt_df)
}

driver_muts$purity = manifest[driver_muts$sample,]$purity
driver_muts$total_cn = c(7, 5, 3, 3, 3, 3, 3, 6, 7) #taken from major clone in Battenberg output
driver_muts$maj_cn = c(4, 4, 3, 3, 3, 3, 3, 4, 5) #taken from major clone in Battenberg output
driver_muts$min_cn = driver_muts$total_cn - driver_muts$maj_cn
driver_muts$mut_multiplicity = driver_muts$VAF * ((driver_muts$total_cn * driver_muts$purity) + (2 * (1 - driver_muts$purity)))/ driver_muts$purity #all consistent with having occurred after WGD

```


#Compare truncal mutations against total tumour burden

```{r}

library(reshape2)
library(ggplot2)
library(ggpubr)

sub_prop = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/gct_trunk_v_overall_subs.txt', header = T, sep = '\t', stringsAsFactors = F, row.names = 1)
sub_prop$patient = row.names(sub_prop)

sub_prop.m = reshape2::melt(sub_prop[, c('Truncal.Overall', 'X1...Truncal.Overall.', 'patient')], id.vars = 'patient')
sub_prop.m$variable = factor(sub_prop.m$variable, levels = c('X1...Truncal.Overall.', 'Truncal.Overall'))
sub_prop.m$patient = factor(sub_prop.m$patient, levels = c("PD46969", "PD46966", "PD46271", "PD46270", "PD46269", "PD45545", "PD45544", "PD45543", "PD43298", "PD43296", "PD42036", "PD43299"))

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_sub_prop_trunk_v_overall_20210304.pdf', height = 4.5, width = 3, useDingbats = F)
ggplot(sub_prop.m) + geom_bar(mapping = aes(x = value, y = patient, fill = variable), position="stack", stat="identity") + scale_fill_manual(values = c(NA, '#999999')) + coord_cartesian(expand = F) + ylab('') + xlab('Proportion of substitutions in phylogenetic trunk') + theme_pubr(legend = 'none') + geom_vline(xintercept = median(sub_prop.m[sub_prop.m$variable == 'Truncal.Overall', ]$value), lty = 2) #0.9191559 median
dev.off()

```

#WGD estimation analysis

Function from PCAWG paper

```{r}

source('/nfs/users/nfs_t/to3/scripts/additional_mutationtimeR_functions.R') #source extra, unchanged functions

#edited functions from mutationtimeR
findMainCluster <- function(bb, min.dist = 0.05){ #at least 20 SNVs
    w <- which(bb$n.snv_mnv >= 20 & !is.na(bb$time) & ((bb$time.up - bb$time.lo) < 0.5)) #edited to exclude regions with very wide confidence intervals when, to3 03/03/21 estimating WGD time
    s <- seq(0,1,0.001) #increased granularity in view of how early WGD is, to3 03/03/21
    l2 <- pmin(bb$time.lo, bb$time - min.dist)[w]
    u2 <- pmax(bb$time.up, bb$time + min.dist)[w]
    l1 <- (l2 +  bb$time[w])/2
    u1 <- (u2 +  bb$time[w])/2
    wd <- as.numeric(width(bb)[w])
    o <- sapply(s, function(i) sum(wd * ( (l2 <= i & u2 >=i) + (l1 <= i & u1 >= i))))
    s[which.max(o)]
}

#remove the deamination component as we have too few mutations, especially associated with SBS1 to do this and filter low confidence timing segments from analysis
computeWgdParam <- function(vcf, bb, clusters, purity, sex){ #added sex in
    # 1. Find segments compatible with WGD
    min.dist <- 0.05
    m <- findMainCluster(bb) #modifed as above to higher confidence regions on which to time WGD, to3 03/03/21
    l <- pmin(bb$time.lo, bb$time - min.dist)
    u <- pmax(bb$time.up, bb$time + min.dist)
    o <- which(l <= m & u >= m & ((bb$time.up - bb$time.lo) < 0.5)) #only going to count and time over segments with higher confidence, to3 03/03/21
    #note that this line has been left as is in fractionGenomeWgdCompatible as there we only want to know the regions compatable with WGD, not necessarily depend on them all for downstream analyses, , to3 03/03/21
    # 2. Find substitutions in compatible segments
    intersect_seg = findOverlaps(vcf, bb[o], ignore.strand = T) #%over% method not working so this is a workaround, to3 03/03/21
    vcf <- vcf[queryHits(intersect_seg),] 
    w <- which(info(vcf)$MajCN==2 & sapply(info(vcf)$CNID, length)==1)
    v <- vcf[w]
    if(nrow(v)< 100) return(NULL) # At least 100 SNVs total over which to estimate the time of acquisition
    seqnames(rowRanges(v)) <- factor(3-info(v)$MinCN, levels=seqlevels(v))
    v <- sort(v)
    
    # 3. Merged CN segments
    b <- GRanges(1:3, IRanges(rep(1,3),rep(max(end(v)),3)), copy_number=4:2, major_cn=2, minor_cn=2:0, clonal_frequency=as.numeric(purity))
    
    # 4. Calculate times
    l <- computeMutCn(v, b, clusters, purity, isWgd=TRUE, n.boot=200, rho=0.01, xmin=3, gender = sex) #recalculates times calculated before on single segments
    b$n.snv_mnv <- l$n <- table(factor(info(v)$MinCN, levels=2:0))
    l$time <- bbToTime(b, l$P)

    row.names(l$D) = row.names(info(v)) #get sub annotations back, to3 03/03/21
    l$cn_state_adj_burden = sum(l$D$MutCN == 2, na.rm = T) + (sum(l$D$MutCN == 2, na.rm = T) * ((unname(l$n['0']) +  unname(l$n['1'])) /  (unname(l$n['0']) +  unname(l$n['1']) + unname(l$n['2'])))) #correct for the 2+1 and 2+0 regions where pre-duplication mutations on the minor allele cannot be counted, to3 04/03/21
    l$hq_seg_width = sum(width(bb[o][bb[o]$major_cn == 2])) #over what length of the genome were these substitutions identified? to3 03/03/21
    l$adj_genome_sub_burden = l$cn_state_adj_burden * (genome_length / l$hq_seg_width) #number of mutations estimated to be on both copies, corrected for genome coverage by good confidence WGD segments, to3 04/03/21
         
    return(l)
}

```

Create lists of pertinent input data, including output files from mutationtimeR

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/mutationtimeR')

chrom_sizes <- read.table('/lustre/scratch119/casm/team294rr/rs30/Testicular_project/Final_Analyses/metadata/male.hg19.chrom.sizes.tsv', header = F, sep = '\t', stringsAsFactors = F)
genome_length = sum(chrom_sizes[1:24,]$V2) #3095677412

bb_files = list.files(pattern = '_post_mutationtime_segments.rds')

finalBB <- list()
for( f in bb_files){
    g = readRDS(f)
    finalBB[[unlist(strsplit(f, '_post_mutationtime_segments.rds'))]] <- g
}

snv_files = list.files(pattern = '_ssm_post_mutationtime.vcf')

finalSnv = list()
for(f in snv_files){
  vcf = readVcf(f)
  finalSnv[[unlist(strsplit(f, '_ssm_post_mutationtime.vcf'))]] <- vcf
}

cluster_files = list.files(pattern = '_10000iters_3000burnin_bestClusterInfo.txt')

finalClusters <- list()
for(f in cluster_files){
  clusters <- read.table(f, sep = '\t', header = T, stringsAsFactors = F)
  sample = unlist(strsplit(f, '_10000iters_3000burnin_bestClusterInfo.txt'))
  bb = read.table(paste0(sample, '_post_mutationtime_segments.txt'), sep = '\t', header = T, stringsAsFactors = F)
  clusters$proportion = clusters$location * max(bb$clonal_frequency)
  clusters <- clusters[, c('cluster.no', 'proportion', 'no.of.mutations')]
  names(clusters) <- c('cluster', 'proportion', 'n_ssms')
  finalClusters[[sample]] <- clusters
}

purity_files = list.files(pattern = '.battenberg.rho_and_psi_updated.txt')

finalPurity <- numeric()
for(f in purity_files){
  pur_df = read.table(f, sep = '\t', header = T, stringsAsFactors = F)
  purity = pur_df['FRAC_GENOME', 'rho']
  finalPurity[unlist(strsplit(f, '.battenberg.rho_and_psi_updated.txt'))] = purity
}

sex_vec = ifelse(grepl('PD455', purity_files), 'female', 'male')
names(sex_vec) = unlist(strsplit(purity_files, '.battenberg.rho_and_psi_updated.txt'))

```

Sanity check for final, eligible samples - are they consistent with WGD?

```{r}

finalPloidy <- sapply(finalBB, averagePloidy)
names(finalPloidy) <- names(finalBB)

finalHom <- sapply(finalBB, averageHom)
names(finalHom) <- names(finalBB)

isWgd <- .classWgd(finalPloidy, finalHom)
table(isWgd) #all samples classed as WGD based on ploidy and homozygosity

fracGenomeWgdComp <- t(sapply(finalBB, function(bb) {
                    fgw <- try(fractionGenomeWgdCompatible(bb)); 
                    if(class(fgw)!='try-error') fgw
                    else rep(NA,10)}))
rownames(fracGenomeWgdComp) <- names(finalBB)
fracGenomeWgdComp = data.frame(fracGenomeWgdComp)
fracGenomeWgdComp$patient = substr(row.names(fracGenomeWgdComp), 0, 7)
aggregate(time.wgd ~ patient, fracGenomeWgdComp, median)

#   patient time.wgd
#  PD42036   0.0220
#  PD43296   0.0155
#  PD43298   0.0570
#  PD43299   0.3935
#  PD45543   0.0455
#  PD45544   0.0290
#  PD45545   0.0195
#  PD46269   0.0100
#  PD46271   0.0175
#  PD46966   0.0380
#  PD46969   0.0065

```

Generate parameters for estimating the pre-duplication mutation burden.

```{r}

wgdParam <- mclapply(names(finalSnv)[isWgd], function(ID){
            try(computeWgdParam(finalSnv[[ID]], finalBB[[ID]], clusters=finalClusters[[ID]], purity=finalPurity[ID], sex=sex_vec[ID]))
        })
names(wgdParam) <- names(finalSnv)[isWgd]
void <- sapply(wgdParam, function(x) is.null(x) | class(x)=="try-error") #none

saveRDS(wgdParam, 'mutationtimeR_WGD_segments_20210303.rds')

predup_burden_df = data.frame(sample = names(wgdParam))
row.names(predup_burden_df) = predup_burden_df$sample

predup_burden_df$adj_genome_sub_burden = NA
predup_burden_df$hq_seg_width = NA

for(i in 1:nrow(predup_burden_df)){
  predup_burden_df$adj_genome_sub_burden[i] = wgdParam[[predup_burden_df$sample[i]]]$adj_genome_sub_burden
  predup_burden_df$hq_seg_width[i] = wgdParam[[predup_burden_df$sample[i]]]$hq_seg_width
}

predup_burden_df$patient = substr(predup_burden_df$sample, 0, 7)
predup_burden_df_per_patient = aggregate(adj_genome_sub_burden ~ patient, predup_burden_df, median)

#   patient adj_genome_sub_burden
#  PD42036              0.000000
#  PD43296              7.151071
#  PD43298             13.736280
#  PD43299            419.612430
#  PD45543              0.000000
#  PD45544              0.000000
#  PD45545              7.617121
#  PD46269              0.000000
#  PD46271              5.903524
#  PD46966              7.210443
#  PD46969             15.341817

median(predup_burden_df_per_patient[predup_burden_df_per_patient$patient != 'PD43299',]$adj_genome_sub_burden) #6.527298

write.table(fracGenomeWgdComp, 'gct_wgd_estimate_mut_time_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)
write.table(predup_burden_df, 'gct_predup_sub_burden_estimate_20210304.txt', col.names = T, row.names = F, sep = '\t', quote = F)

```

```{r}

fracGenomeWgdComp = read.table('gct_wgd_estimate_mut_time_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)

library(reshape2)
library(ggplot2)
library(ggpubr)
library(VariantAnnotation)
library(ggbeeswarm)

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/wgd_timing_per_microbiopsy_20210309.pdf', useDingbats = F, height = 6, width = 6)
ggplot(fracGenomeWgdComp) +
  geom_quasirandom(mapping = aes(x = patient, y = time.wgd), col = '#999999') +
  geom_pointrange(mapping = aes(x = patient, y = time.wgd),
                  stat = "summary", shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill="black") +
  theme_pubr() + ylab('Estimated WGD timing') + xlab('') +
  scale_x_discrete(expand = c(0,0.5)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0.01)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust = 0.9))
dev.off()

table(fracGenomeWgdComp$patient)
#PD42036 PD43296 PD43298 PD43299 PD45543 PD45544 PD45545 PD46269 PD46271 PD46966 PD46969 
#      5      22      20       4       2       5       2       3       2       3       4 

```

##Plot pre-WGD sub burden against truncal mutation burden

```{r}

library(reshape2)
library(ggplot2)
library(ggpubr)
library(VariantAnnotation)
library(ggbeeswarm)

sub_prop = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/gct_trunk_v_overall_subs.txt', header = T, sep = '\t', stringsAsFactors = F, row.names = 1)
sub_prop$patient = row.names(sub_prop)

sub_prop.m = reshape2::melt(sub_prop[, c('Truncal.Overall', 'X1...Truncal.Overall.', 'patient')], id.vars = 'patient')
sub_prop.m$variable = factor(sub_prop.m$variable, levels = c('X1...Truncal.Overall.', 'Truncal.Overall'))
sub_prop.m$patient = factor(sub_prop.m$patient, levels = c("PD46969", "PD46966", "PD46271", "PD46270", "PD46269", "PD45545", "PD45544", "PD45543", "PD43298", "PD43296", "PD42036", "PD43299"))

timing_tab = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/mutationtimeR/gct_predup_sub_burden_estimate_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/pre_dup_burden_per_microbiopsy_20210304.pdf', useDingbats = F, height = 6, width = 6)
ggplot(timing_tab) +
  geom_quasirandom(mapping = aes(x = patient, y = adj_genome_sub_burden), col = '#999999') +
  geom_pointrange(mapping = aes(x = patient, y = adj_genome_sub_burden),
                  stat = "summary", shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill="black") +
  theme_pubr() + ylab('Estimated pre-duplication substution burden') + xlab('') +
  scale_x_discrete(expand = c(0,0.5)) +
  scale_y_continuous(limits = c(0, 500), expand = c(0, 0.1)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust = 0.9))
dev.off()

final_wgd_plotting_agg = aggregate(adj_genome_sub_burden ~ patient, timing_tab, median)

final_wgd_plotting_agg$trunc_mt = NA
for(i in 1:nrow(final_wgd_plotting_agg)){
  final_wgd_plotting_agg$trunc_mt[i] = sub_prop[final_wgd_plotting_agg$patient[i], ]$Invasive.truncal.sum
}

median(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$adj_genome_sub_burden) #6.527298
final_wgd_plotting_agg$pre_WGD_burden_log = log(final_wgd_plotting_agg$adj_genome_sub_burden + 1) #to keep all values over 0
final_wgd_plotting_agg$trunc_mt_log = log(final_wgd_plotting_agg$trunc_mt + 1)

final_wgd_plotting_agg.m = reshape2::melt(final_wgd_plotting_agg[, c('patient', 'pre_WGD_burden_log', 'trunc_mt_log')], id.vars = 'patient')

final_wgd_plotting_agg.m$Patient = factor(final_wgd_plotting_agg.m$patient, levels = c("PD46969", "PD46966", "PD46271", "PD46270", "PD46269", "PD45545", "PD45544", "PD45543", "PD43298", "PD43296", "PD42036", "PD43299"))
final_wgd_plotting_agg.m$variable = factor(final_wgd_plotting_agg.m$variable, levels = c('trunc_mt_log', 'pre_WGD_burden_log'))

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_sub_prop_predup_v_trunc_mt_20210304.pdf', height = 4.5, width = 3, useDingbats = F)
ggplot(final_wgd_plotting_agg.m) +
  geom_bar(mapping = aes(x = value, y = Patient, fill = variable), position="dodge", stat="identity") + scale_fill_manual(values = c('grey80', 'grey40')) + coord_cartesian(expand = F, xlim = c(log(0 + 1), log(10000 + 1))) + ylab('') + xlab('Number of substitutions') + theme_pubr(legend = 'none') + scale_x_continuous(breaks = c(log(0 + 1), log(10 + 1), log(100 + 1), log(1000 + 1), log(10000 + 1)), labels = c(0, 10, 100, 1000, 10000)) + geom_vline(xintercept = median(final_wgd_plotting_agg.m[final_wgd_plotting_agg.m$variable == 'pre_WGD_burden_log' & final_wgd_plotting_agg.m$patient != 'PD43299', ]$value), lty = 2) #2.015091, actual value underlying this is 6.527298
dev.off()
```

##Estimate the number of cell divisions prior to WGD


```{r}

final_wgd_plotting_agg$estimated_postPGC_cell_divisions = final_wgd_plotting_agg$adj_genome_sub_burden / 0.6 #middle of range for estimated subs per cell division

#take estimate of postPGC mutation rate per cell division in testis and ovary from Rahbari et al. 2016
final_wgd_plotting_agg$estimated_postPGC_cell_divisions_upper = final_wgd_plotting_agg$adj_genome_sub_burden / 0.5
final_wgd_plotting_agg$estimated_postPGC_cell_divisions_lower = final_wgd_plotting_agg$adj_genome_sub_burden / 0.7

median(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$estimated_postPGC_cell_divisions) #10.87883
median(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$estimated_postPGC_cell_divisions_lower) #9.324711
median(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$estimated_postPGC_cell_divisions_upper) #13.0546

min(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$estimated_postPGC_cell_divisions_lower) #0
max(final_wgd_plotting_agg[final_wgd_plotting_agg$patient != 'PD43299',]$estimated_postPGC_cell_divisions_upper) #30.68363

```

##Compare WGD timing to PCAWG

PCAWG data kindly provided by Peter van Loo

```{r}

summary_data = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_DNA_supplementary_data.txt', header = T, sep = '\t', stringsAsFactors = F)
row.names(summary_data) = summary_data$Sample

#read in PCAWG data
pcawg_wgd = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/20190824_timing_gains_adj.txt', header = T, sep = '\t', stringsAsFactors = F)

pcawg_wgd[pcawg_wgd$samplename %in% pcawg_wgd[pcawg_wgd$type == 'WGD',]$samplename,]


sample_hist_key = pcawg_wgd[!duplicated(pcawg_wgd[, c('samplename', 'histology_abbreviation')]), c('samplename', 'histology_abbreviation')]
pcawg_wgd = pcawg_wgd[pcawg_wgd$histology_abbreviation %in% sort(unique(sample_hist_key$histology_abbreviation))[table(sample_hist_key$histology_abbreviation) >= 10], ] #only keep tumour types with at least 10 samples to make proportion values sensible

sample_hist_key = sample_hist_key[sample_hist_key$histology_abbreviation %in% sort(unique(sample_hist_key$histology_abbreviation))[table(sample_hist_key$histology_abbreviation) >= 10], ]

sample_hist_key$WGD = NA
sample_hist_key$WGD_timing = NA

for(i in 1:nrow(sample_hist_key)){
  sample_hist_key$WGD[i] = 'No'
  if(sample_hist_key$samplename[i] %in% pcawg_wgd[pcawg_wgd$chr == 'WGD',]$samplename){
    sample_hist_key$WGD[i] = 'Yes'
    sample_hist_key$WGD_timing[i] = pcawg_wgd[pcawg_wgd$samplename == sample_hist_key$samplename[i] & pcawg_wgd$chr == 'WGD',]$time
  } 
}

#read in GCT data
timing_df = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/mutationtimeR/gct_wgd_estimate_mut_time_20210304.txt', sep = '\t', header = T, stringsAsFactors = F)

median_wgd = aggregate(time.wgd ~ patient, timing_df, median)
median_wgd_postpubertal = median_wgd[median_wgd$patient != 'PD43299',]
median_wgd_postpubertal$histology_abbreviation = 'Postpubertal_GCT'

gct_samples = data.frame(samplename = median_wgd_postpubertal$patient, histology_abbreviation = rep('Postpubertal_GCT', nrow(median_wgd_postpubertal)), WGD = rep('Yes', nrow(median_wgd_postpubertal)), WGD_timing = median_wgd_postpubertal$time.wgd)

median(median_wgd_postpubertal[median_wgd_postpubertal$patient %in% c('PD45543', 'PD45544', 'PD45545'),]$time.wgd) #post-pubertal ovary 0.029
median(median_wgd_postpubertal[!median_wgd_postpubertal$patient %in% c('PD45543', 'PD45544', 'PD45545'),]$time.wgd) #post-pubertal testis 0.0175

#combined data gct and pcawg data
sample_hist_key = rbind(sample_hist_key, gct_samples)

sample_hist_key$WGD_timing_plot = round(sample_hist_key$WGD_timing, digits = 2)
sample_hist_key[is.na(sample_hist_key$WGD_timing_plot),]$WGD_timing_plot = 'None'
sample_hist_key$WGD_timing_plot = factor(sample_hist_key$WGD_timing_plot, levels = rev(sort(unique(sample_hist_key$WGD_timing_plot))))

#rank by proportion of samples that undergo WGD and then arrange by the estimated time at which it occurs
ranked_histo = as.data.frame(unique(sample_hist_key$histology_abbreviation))
names(ranked_histo) = 'Dx'

ranked_histo$noWGD_count = NA
ranked_histo$total = NA

for(i in 1:nrow(ranked_histo)){
  ranked_histo$noWGD_count[i] = nrow(sample_hist_key[sample_hist_key$histology_abbreviation == ranked_histo$Dx[i] & sample_hist_key$WGD == 'No',])
  ranked_histo$total[i] = nrow(sample_hist_key[sample_hist_key$histology_abbreviation == ranked_histo$Dx[i],])
}

ranked_histo$prop_noWGD = ranked_histo$noWGD_count / ranked_histo$total
sample_hist_key$histology_abbreviation = factor(sample_hist_key$histology_abbreviation, levels = ranked_histo[order(ranked_histo$prop_noWGD, decreasing = T),]$Dx)

#g <- colorRampPalette(RColorBrewer::brewer.pal(4,"Set1")[c(3,2,4)])(101)
g <- colorRampPalette(c('grey80', 'black'))(101)
names(g) = seq(0.0, 1, 0.01)
h = c(g, 'white')
names(h)[102] = 'None'

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/wgd_gct_v_pcawg_plot_20210304.pdf', height = 6, width = 8)
ggplot(sample_hist_key, aes(x = histology_abbreviation, fill = WGD_timing_plot)) +
  geom_bar(position="fill") + 
  scale_fill_manual(values = h) + 
  coord_cartesian(expand = F) +
  theme_classic() +
  theme(legend.position = 'none', axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2)) +
  xlab('')  + ylab('Proportion undergone WGD')
dev.off()

```

#TCGA chr12 configurations

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data')

#get metadata, including ploidy values for all eligible tumours
tcga_ploidy = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct/summary.ascatTCGA.penalty70.hg19.tsv', header = T, sep = '\t', stringsAsFactors = F)
tcga_ploidy = tcga_ploidy[(tcga_ploidy$cancer_type == 'TGCT') & (tcga_ploidy$pass == TRUE) & (tcga_ploidy$rep == TRUE) & (tcga_ploidy$purity > 0.4), ] #103 samples

#get copy segments for all tumours in a single dataframe
tcga_segment_files = list.files(path = '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct', pattern = 'segments.txt')

tcga_ascat_segments = c()
for(i in tcga_ploidy$name){
  tumour_file = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/tcga_tgct/', i, '.segments.txt'), header = T, sep = '\t', stringsAsFactors = F)
  tcga_ascat_segments = rbind(tcga_ascat_segments, tumour_file)
}

names(tcga_ascat_segments) = c('tumour', 'chr', 'start', 'end', 'majA', 'minA')
tcga_ascat_segments$ntot = tcga_ascat_segments$majA + tcga_ascat_segments$minA

tcga_ascat_chr12_segments = tcga_ascat_segments[tcga_ascat_segments$chr == 12,]

#remove all tumours where any region of 12p has a coopy number greater than 4
write.table(tcga_ascat_chr12_segments[!(tcga_ascat_chr12_segments$tumour %in% (unique(tcga_ascat_chr12_segments[(tcga_ascat_chr12_segments$start < 35800000 & tcga_ascat_chr12_segments$ntot > 4),]$tumour))),], 'tcga_chr12p_low_gain_segments_20210306.txt', col.names = T, row.names = F, sep = '\t', quote = F)

#7 cases show no 12p gains beyond that achieved during WGD.

```

#RNA filtering & pre-processing
##Filtering
Raw count table, preQC with genes mapped to nuclear genome only.

```{r}

library(ggplot2)
library(ggpubr)

raw_data = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/filtering/germ_cell_lcm_rna_preQC.txt', header = T, sep = '\t', stringsAsFactors = F) #just normal testis and GCT raw count data, 500 microbiopsies

#remove transcriptomes that were:
#- identified as contaminated during microdissection
#- non-neoplastic tissues that are not healthy seminiferous tubules
#- uncertainty as to the histological categorisation of the microbiopsy

raw_data = raw_data[, !colnames(raw_data) %in% c("PD43298a_Z1_2_SL02_04_SEC02_04_F2", 'PR43299c_lo0009', 'PR43299c_lo0010', 'PR43299c_lo0011', 'PR43299c_lo0012', "PD43296a_SL02_SEC02_G1", "PD43296a_Z1_2_SL02_04_SEC02_04_E1A", "PD43298a_SL02_SEC02_B3", "PD43298a_SL02_SEC02_C3", "PD43298a_Z1_2_SL02_04_SEC02_04_G2", "PR43299c_lo0002", "PR43299c_lo0006", "PR42036a_SL0D_SEC02_A5", "PR42036a_SL0D_SEC02_G3", "PR42036a_SL0D_SEC02_H3", "PR43298a_SL04_SEC04_D5",             "PR43298a_SL04_SEC04_E5" ,"PR43298a_SL04_SEC04_G9", "PR45543a_SL02_SEC03_B4", "PR45543a_SL02_SEC03_D2", "PR45543a_SL02_SEC03_F2", "PR46269c_SL02_SEC03_E12", "PR46269c_SL02_SEC03_F12", "PR46270c_SL02_SEC03_B5", "PR46270c_SL02_SEC03_C5", "PR46270c_SL02_SEC03_D5", "PR46270c_SL02_SEC03_E3", "PR46270c_SL02_SEC03_F3",             "PR46270c_SL02_SEC03_G3", "PR46270c_SL02_SEC03_H3", "PR46967a_SL02_SEC03_D8", "PR46967a_SL02_SEC03_E4", "PR46967a_SL02_SEC03_G6", "PR46968d_SL02_SEC02_03_H5", "PR46969c_SL02_SEC03_C11", "PR46969c_SL02_SEC03_D11", "PR46969c_SL02_SEC03_E11", "PR46969c_SL02_SEC03_F11", "PR46969c_SL02_SEC03_G11", "PR46969c_SL02_SEC03_H11")] #460 samples left

#how many reads, across all 55502 mapped features do we have?
summary(colSums(raw_data[, 7:ncol(raw_data)])) #highly variable library sizes
#   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   682   86612  246980  502107  651605 3863511 

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/all_samples_read_count_QC.pdf', height = 6, width = 6, useDingbats = F)
ggplot() +
  geom_histogram(mapping = aes(colSums(raw_data[, 7:ncol(raw_data)])), bins = 100, fill = '#999999') + theme_pubr() + coord_cartesian(expand = F) + xlab('Number of reads across all mapped features') + ylab('Number of microbiopsies')
#spike in samples with under 1000 features expressed
dev.off()

#if we choose a threshold of 5 reads as the minimum to consider a gene truly expressed, how many are expressed per microbiopsy?
summary(colSums(raw_data[, 7:ncol(raw_data)] >= 5))
#Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#   0    4150    8916    9077   13654   22865 

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/all_samples_feature_expression_QC.pdf', height = 6, width = 6, useDingbats = F)
ggplot() +
  geom_histogram(mapping = aes(colSums(raw_data[, 7:ncol(raw_data)] >= 5)), bins = 100, fill = '#999999') + theme_pubr() + coord_cartesian(expand = F) + geom_vline(xintercept = 1000, lty = 2, col = 'red') + xlab('Features expressed per microbiopsy to a depth of at least 5 reads') + ylim(0, 15) + ylab('Number of microbiopsies')
#spike in samples with under 1000 features expressed
dev.off()

```

##Pre-processing 

```{r}

library(limma)
library(edgeR)
library(Seurat)

raw_data = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/filtering/germ_cell_lcm_rna_preQC.txt', header = T, sep = '\t', stringsAsFactors = F) #raw count data
metadata = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_RNA_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) #read in metadata for filtered, final cuts
row.names(metadata) = metadata$Sample_ID
gct_mat = raw_data[, 7:ncol(raw_data)]
row.names(gct_mat) = raw_data$Geneid

#remove cuts that failed QC
gct_mat = gct_mat[, colnames(gct_mat) %in% metadata$Sample_ID] #416 samples

#define factors that will be input into the model
histology = factor(metadata[colnames(gct_mat), ]$Updated.Description)
patient = factor(metadata[colnames(gct_mat), ]$Case_ID)
metadata = metadata[colnames(gct_mat),]

#preprocessing
y <- DGEList(counts = gct_mat, group = histology)
keep <- filterByExpr.default(y, group = histology)
keep <- filterByExpr(y, group = histology)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y) #33185 genes kept
y <- estimateCommonDisp(y) #to generate pseudocounts, adjusted for library size
pseudo_counts = y$pseudo.counts

#check that the tmm-normalised counts cluster by histology before performing DE analysis
gct_obj <- CreateSeuratObject(counts = y$pseudo.counts, meta.data = metadata, project = 'gct')
gct_obj <- NormalizeData(gct_obj, normalization.method = "LogNormalize", scale.factor = 10000)
gct_obj <- FindVariableFeatures(gct_obj, selection.method = "vst", nfeatures = 1000)
all.genes <- rownames(gct_obj )
gct_obj  <- ScaleData(gct_obj , features = all.genes)
gct_obj <- RunPCA(gct_obj, features = VariableFeatures(object = gct_obj), npcs = 100)
DimPlot(gct_obj, reduction = "pca", group.by = 'Updated.Description')
ElbowPlot(gct_obj, ndims = 100)
gct_obj <- RunUMAP(gct_obj, dims = 1:30)

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_seurat_30_dims_histology_umap_20210304.pdf', width = 8, height = 6, useDingbats = F)
DimPlot(gct_obj, reduction = "umap", group.by = 'Updated.Description')
dev.off()

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_seurat_30_dims_patient_umap_20210304.pdf', width = 8, height = 6, useDingbats = F)
DimPlot(gct_obj, reduction = "umap", group.by = 'Case_ID')
dev.off()

saveRDS(gct_obj, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/lcm_gct_seurat_object_20210403.rds')

#clusters by histology but also clearly some patient-specific effect which will need to be accounted for in the differential expression analysis

#save filtered raw counts and tmm-normalised counts, plus calculate tpm values
write.table(pseudo_counts, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/lcm_rna_tmm_norm_counts_20210304.txt', quote = F, col.names = T, row.names = T, sep = '\t')

write.table(y$counts, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/lcm_rna_raw_counts_20210304.txt', quote = F, col.names = T, row.names = T, sep = '\t')

row.names(raw_data) = raw_data$Geneid
edgeR_gene_data = raw_data[row.names(pseudo_counts),]
x = pseudo_counts
x = x / edgeR_gene_data$Length #normalise for transcript length
tpm <- t( t(x) * 1e6 / colSums(x)) #normalise to read depth
write.table(tpm, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/tmm_norm_gct_tpm_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

metadata$total_reads_filtered_genes = NA

for(i in 1:nrow(metadata)){
  metadata$total_reads_filtered_genes[i] = sum(y$counts[, row.names(metadata)[i]])
}

write.table(metadata, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_RNA_metadata.txt', quote = F, col.names = T, row.names = F, sep = '\t')

```

#Limma-Voom DE analysis between different LCM tissues

Continue with objects from above

```{r}

#specify model to be fitted, no intercept to be fitted, each histology will have its own group mean
design = model.matrix(~ 0 + histology)
colnames(design) = levels(histology)

#run double voom to optimise regression weights, adjust for random effect of patient as 'block', may take a couple of hours
v <- voom(y, design)
corfit <- duplicateCorrelation(v, design, block = patient)
v <- voom(y, design, block = patient, correlation = corfit$consensus)
corfit <- duplicateCorrelation(v, design, block = patient)
fit <- lmFit(v, design, block = patient, correlation = corfit$consensus)
corfit$consensus #0.3086162

saveRDS(v, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de/limma_voom_de_LCM_GCT_DE_voom_object_20210304.RDS')
saveRDS(fit, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de/limma_voom_de_LCM_GCT_DE_model_fit_20210304.RDS')

```

##Against seminiferous tubules

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de/')

#General comparison against normal seminiferous tubules, EC and between some other components
##EC
cont_matrix = makeContrasts(Embryonal_carcinoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'ec_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##SEM
cont_matrix = makeContrasts(Seminoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'sem_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Adipose teratoma
cont_matrix = makeContrasts(Adipose_teratoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'fat_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Cartilage teratoma
cont_matrix = makeContrasts(Immature_cartilage_teratoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'cartilage_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Dysgerminoma
cont_matrix = makeContrasts(Dysgerminoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'dysgerminoma_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##GCNIS
cont_matrix = makeContrasts(GCNIS - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'gcnis_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Glandular_teratoma
cont_matrix = makeContrasts(Mature_glandular_teratoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'glandter_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Neuroepithelium
cont_matrix = makeContrasts(Neuroepithelium - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'neuroepi_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Smooth_muscle_teratoma
cont_matrix = makeContrasts(Smooth_muscle_teratoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'smoothmusc_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##other immature/indeterminate maturity_teratoma
cont_matrix = makeContrasts(Other_epithelial_teratoma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'other_ter_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Stroma
cont_matrix = makeContrasts(Stroma - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'stroma_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Syncytiotrophoblasts
cont_matrix = makeContrasts(Syncytiotrophoblasts - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'stb_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

##Yolk_sac_tumour
cont_matrix = makeContrasts(Yolk_sac_tumour - Seminiferous_tubule, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'yst_normtestis_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

```

##Against embryonal carcinoma

```{r}

#Fat vs EC
cont_matrix = makeContrasts(Adipose_teratoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'fat_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Cartilage vs EC
cont_matrix = makeContrasts(Immature_cartilage_teratoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'cartilage_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#GCNIS vs EC
cont_matrix = makeContrasts(GCNIS - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'gcnis_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Glandular_teratoma vs EC
cont_matrix = makeContrasts(Mature_glandular_teratoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'glandter_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Immature epi_teratoma vs EC
cont_matrix = makeContrasts(Other_epithelial_teratoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'otherter_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Neuroepithelium vs EC
cont_matrix = makeContrasts(Neuroepithelium - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'neuroepi_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#SEM vs EC
cont_matrix = makeContrasts(Seminoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'sem_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Smooth_muscle_teratoma vs EC
cont_matrix = makeContrasts(Smooth_muscle_teratoma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'smoothmusc_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Stroma vs EC
cont_matrix = makeContrasts(Stroma - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'stroma_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Syncytiotrophoblasts vs EC
cont_matrix = makeContrasts(Syncytiotrophoblasts - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'stb_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#Yolk_sac_tumour vs EC
cont_matrix = makeContrasts(Yolk_sac_tumour - Embryonal_carcinoma, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(y$counts), sort.by = "P")
head(top.table, 20)
write.table(top.table, 'yst_ec_voom_de_20210304.txt', col.names = T, row.names = T, sep = '\t', quote = F)

```

#Comparing expression of fetal and lineage-specific genes between GCTs and reference single cell data

Using tpm values

```{r}

library(ggplot2)
library(reshape2)
library(cowplot)
library(Seurat)
library(ggbeeswarm)
library(dplyr)
library(ggpubr)
library(gridExtra)

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets')

#read in scRNA seurat objects containing the count data
testis_ref = readRDS('postnatal_testis_final.rds') #neonatal and adult testis - Sohni et al.
fetal_brain = readRDS('pfc_final.rds') #fetal brain tissues - Zhong et al.
adult_brain = readRDS('adult_brain_li.rds') #adult brain - Li et al.
fetal_muscle = readRDS('fetal_muscle_cao.rds') #fetal cardiac, smooth and skeletal muscle - Cao et al.
adult_smooth_musc = readRDS('adult_smooth_muscle.rds') #adult cardiac, smooth and skeletal muscle - Litviukov et al.

#read in our TPM data
gct_tpm = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/tmm_norm_gct_tpm_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)

#merge reference scRNA data and restrict to genes also found in our data
ref_scrna = merge(testis_ref, y = c(fetal_brain, adult_brain, fetal_muscle, adult_smooth_musc), add.cell.ids = c("testis", "fetal_brain", "adult_brain","fetal_smooth_muscle", "adult_smooth_muscle"), project = 'ref_scrna_data') #49128 features across 32036 samples within 1 assay 

genes_to_keep = row.names(ref_scrna)[row.names(ref_scrna) %in% row.names(pseudo_counts)]

ref_scrna_final = CreateSeuratObject(counts = ref_scrna@assays$RNA@counts[genes_to_keep,], project = 'final_ref_scrna', assay = 'RNA', meta.data = ref_scrna@meta.data) #27569 features across 46195 samples within 1 assay

ref_scrna_final = subset(ref_scrna_final, subset = cell_types != 'Microglia' & cell_types != 'Pre-Spermatogonia') #27168 features across 32036 samples within 1 assay

ref_scrna_final = NormalizeData(object = ref_scrna_final, normalization.method = "LogNormalize")
all.genes <- rownames(ref_scrna_final)
ref_scrna_final <- ScaleData(ref_scrna_final, features = all.genes)
ref_scrna_final@meta.data$cell_types = factor(ref_scrna_final@meta.data$cell_types)

levels(ref_scrna_final@meta.data$cell_types) = c("Adult astrocytes", "Fetal astrocytes", "Fetal excitatory neurons", "Adult excitatory neurons", "Fetal cardiac smooth muscle cells", "Adult interneurons", "Fetal interneurons", "Fetal other smooth muscle cells", "Fetal neuronal stem cells", "Adult oligodendrocytes", "Adult oligodendrocyte progenitor cells", "Fetal oligodendrocyte progenitor cells", "PGC-like", "Adult other smooth muscle cells", "Spermatid", "Spermatocytes", "Spermatogonia")

saveRDS(ref_scrna_final, 'reference_scrna.rds')

#ref_scrna_final = readRDS('reference_scrna.rds')

ref_scrna_final@meta.data$cell_types = as.factor(ref_scrna_final@meta.data$cell_types)

ref_scrna_final@meta.data$cell_types = factor(ref_scrna_final@meta.data$cell_types, levels = rev(c("PGC-like", "Spermatogonia", "Spermatocytes", "Spermatid", "Fetal neuronal stem cells", "Fetal oligodendrocyte progenitor cells", "Adult oligodendrocyte progenitor cells", "Adult oligodendrocytes", "Fetal astrocytes", "Adult astrocytes", "Fetal interneurons", "Adult interneurons", "Fetal excitatory neurons", "Adult excitatory neurons", "Fetal cardiac smooth muscle cells", "Fetal other smooth muscle cells", "Adult other smooth muscle cells")))

#limit gct tpm counts to genes shared between datasets
gct_tpm_filtered = gct_tpm[row.names(ref_scrna_final),]
gct_log2tpm <- log2(as.matrix(gct_tpm_filtered) + 1)
rna_cuts = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_RNA_metadata.txt', header = T, sep = '\t', stringsAsFactors = F)
row.names(rna_cuts) = rna_cuts$Sample_ID

#extract ref data counts and convert to tpm
ref_scrna_final_df = as.data.frame(ref_scrna_final@assays$RNA@counts)
write.table(ref_scrna_final_df, 'reference_scrna_counts.txt', col.names = T, row.names = T, sep = '\t', quote = F)

ref_scrna_final_tpm <- t( t(ref_scrna_final_df) * 1e6 / colSums(ref_scrna_final_df)) #tpm conversion, 
ref_scrna_final_log2tpm <- log2(as.matrix(ref_scrna_final_tpm) + 1)

write.table(ref_scrna_final_log2tpm, 'reference_scrna_log2tpm.txt', col.names = T, row.names = T, sep = '\t', quote = F)

#ref_scrna_final_log2tpm = read.table('reference_scrna_log2tpm.txt', header = T, sep = '\t', stringsAsFactors = F)

ref_scrna_final@meta.data$sample = row.names(ref_scrna_final@meta.data)
ref_scrna_final@meta.data$source = 'scrna'
ref_scrna_final_meta = ref_scrna_final@meta.data[, c('sample', 'cell_types', 'source')]

rna_cuts$cell_types = rna_cuts$Updated.Description
rna_cuts$sample = row.names(rna_cuts)
rna_cuts$source = 'lcm'
gct_final_meta = rna_cuts[, c('sample', 'cell_types', 'source')]

all_data_final_log2tpm = cbind(ref_scrna_final_log2tpm, gct_log2tpm)
all_data_final_meta = rbind(ref_scrna_final_meta, gct_final_meta)

colnames(all_data_final_log2tpm) = gsub('\\.', '_', colnames(all_data_final_log2tpm))
row.names(all_data_final_meta) = gsub('-', '_', row.names(all_data_final_meta))
row.names(all_data_final_meta) = gsub('\\.', '_', row.names(all_data_final_meta))

#germ cell comparison
gc_exp = all_data_final_log2tpm[,row.names(all_data_final_meta[all_data_final_meta$cell_types %in% c('PGC-like', 'Spermatogonia', 'Spermatocytes', 'Spermatid', 'GCNIS', 'Seminoma', 'Dysgerminoma', 'Seminiferous_tubule'), ])]
gc_exp.t = as.data.frame(t(as.data.frame(gc_exp)))
gc_exp.t$cell_types = all_data_final_meta[row.names(gc_exp.t),]$cell_types
gc_exp.t$sample = row.names(gc_exp.t)
gc_exp.t.m = reshape2::melt(gc_exp.t[, c('cell_types', 'sample', 'POU5F1', 'NANOG', 'SOX17', 'DDX4', 'TNP1')], id.vars = c('cell_types', 'sample'))

gc_exp.t.m$gene_use = NA
for(i in 1:nrow(gc_exp.t.m)){
  if(gc_exp.t.m$variable[i] %in% c('POU5F1', 'NANOG', 'SOX17')) gc_exp.t.m$gene_use[i] = 'fetal'
  if(gc_exp.t.m$variable[i] == 'DDX4') gc_exp.t.m$gene_use[i] = 'gen_gc'
  if(gc_exp.t.m$variable[i] == 'TNP1') gc_exp.t.m$gene_use[i] = 'spermatid'
}

gc_exp.t.m$gene_use = factor(gc_exp.t.m$gene_use, levels = c('fetal', 'gen_gc', 'spermatid'))
gc_exp.t.m$cell_types = factor(gc_exp.t.m$cell_types, levels = c('PGC-like', 'Spermatogonia', 'Spermatocytes', 'Spermatid', 'GCNIS', 'Seminoma', 'Dysgerminoma', 'Seminiferous_tubule'))

aggregate(value ~ cell_types + variable, gc_exp.t.m, length) #Dysgerminoma will need individual data points for DDX4 and TNP1 plotting

p1 = ggplot(gc_exp.t.m, mapping = aes(x = gene_use, y = value)) + 
  geom_quasirandom(dodge.width=.8, cex=1, aes(x = gene_use, y = value), shape = 21, fill = 'transparent', col = 'transparent') + 
  geom_beeswarm(data = gc_exp.t.m[gc_exp.t.m$cell_types == 'Dysgerminoma' & gc_exp.t.m$variable %in% c('DDX4', 'TNP1'), ], mapping = aes(x = gene_use, y = value), dodge.width=1, cex=1, shape = 21, fill = 'black', col = 'black') + 
  facet_wrap(. ~ cell_types, strip.position = 'bottom', ncol = 8) +
  theme_pubr() + 
  theme(axis.text.x = element_blank(), 
        panel.spacing = unit(0, "mm"), 
        strip.background = element_blank(), 
        strip.placement = "outside", 
        legend.position = 'none', 
        axis.ticks.x = element_blank()) +
  labs(y = 'Log2(TPM) expression', x = '') + 
  geom_pointrange(mapping = aes(col = gene_use), stat = "summary", shape=19, fun.min = function(z) {quantile(z,0.25)}, 
                  fun.max = function(z) {quantile(z,0.75)}, 
                  fun = median, size = 0.8, fatten = 5, linetype = 2) + 
  scale_color_manual(values = c('#FF6C5C', '#58CCED', '#072F5F'))

#neural comparison
ne_exp = all_data_final_log2tpm[,row.names(all_data_final_meta[all_data_final_meta$cell_types %in% c('Neuroepithelium', 'Fetal interneurons', 'Fetal neuronal stem cells', 'Fetal oligodendrocyte progenitor cells', 'Fetal astrocytes', 'Fetal excitatory neurons', 'Adult interneurons', 'Adult oligodendrocyte progenitor cells', 'Adult astrocytes', 'Adult excitatory neurons', 'Adult oligodendrocytes'), ])]
ne_exp.t = as.data.frame(t(as.data.frame(ne_exp)))
ne_exp.t$cell_types = all_data_final_meta[row.names(ne_exp.t),]$cell_types
ne_exp.t$sample = row.names(ne_exp.t)
ne_exp.t.m = reshape2::melt(ne_exp.t[, c('cell_types', 'sample', 'PAX6', 'SOX2', 'NES', 'MBP', 'SLC1A2', 'RBFOX3', 'GAD1')], id.vars = c('cell_types', 'sample'))

ne_exp.t.m$gene_use = NA
for(i in 1:nrow(ne_exp.t.m)){
  if(ne_exp.t.m$variable[i] %in% c('PAX6', 'SOX2', 'NES')) ne_exp.t.m$gene_use[i] = 'fetal'
  if(ne_exp.t.m$variable[i] == 'MBP') ne_exp.t.m$gene_use[i] = 'oligo'
  if(ne_exp.t.m$variable[i] == 'SLC1A2') ne_exp.t.m$gene_use[i] = 'astro'
  if(ne_exp.t.m$variable[i] == 'RBFOX3') ne_exp.t.m$gene_use[i] = 'neun'
  if(ne_exp.t.m$variable[i] == 'GAD1') ne_exp.t.m$gene_use[i] = 'inter'
}

ne_exp.t.m$gene_use = factor(ne_exp.t.m$gene_use, levels = c('fetal', 'oligo', 'astro', 'neun', 'inter'))
ne_exp.t.m$cell_types = factor(ne_exp.t.m$cell_types, levels = c('Fetal neuronal stem cells', 'Fetal oligodendrocyte progenitor cells', 'Fetal astrocytes', 'Fetal excitatory neurons', 'Fetal interneurons', 'Adult oligodendrocyte progenitor cells', 'Adult oligodendrocytes', 'Adult astrocytes', 'Adult excitatory neurons', 'Adult interneurons', 'Neuroepithelium'))

aggregate(value ~ cell_types + variable, ne_exp.t.m, length)

p2 = ggplot(ne_exp.t.m, mapping = aes(x = gene_use, y = value)) + 
  geom_quasirandom(dodge.width=.8, cex=1, aes(x = gene_use, y = value), shape = 21, fill = 'transparent', col = 'transparent') + 
  facet_wrap(. ~ cell_types, strip.position = 'bottom', ncol = 11) +
  theme_pubr() + 
  theme(axis.text.x = element_blank(), 
        panel.spacing = unit(0, "mm"), 
        strip.background = element_blank(), 
        strip.placement = "outside", 
        legend.position = 'none', 
        axis.ticks.x = element_blank()) +
  labs(y = 'Log2(TPM) expression', x = '') + 
  geom_pointrange(mapping = aes(col = gene_use), stat = "summary", shape=19, fun.min = function(z) {quantile(z,0.25)}, 
                  fun.max = function(z) {quantile(z,0.75)}, 
                  fun = median, size = 0.8, fatten = 5, linetype = 2) + 
  scale_color_manual(values = c('#FF6C5C', '#58CCED', '#3895D3', '#1261A0', '#072F5F'))

#smooth muscle comparison
sm_exp = all_data_final_log2tpm[,row.names(all_data_final_meta[all_data_final_meta$cell_types %in% c("Fetal cardiac smooth muscle cells", "Fetal other smooth muscle cells", "Adult other smooth muscle cells", "Smooth_muscle_teratoma"), ])]
sm_exp.t = as.data.frame(t(as.data.frame(sm_exp)))
sm_exp.t$cell_types = all_data_final_meta[row.names(sm_exp.t),]$cell_types
sm_exp.t$sample = row.names(sm_exp.t)
sm_exp.t.m = reshape2::melt(sm_exp.t[, c('cell_types', 'sample', 'IGF2', 'TAGLN', 'ACTA2', 'MYH11')], id.vars = c('cell_types', 'sample'))

sm_exp.t.m$gene_use = NA
for(i in 1:nrow(sm_exp.t.m)){
  if(sm_exp.t.m$variable[i] %in% c('TAGLN', 'ACTA2', 'MYH11')) sm_exp.t.m$gene_use[i] = 'adult'
  if(sm_exp.t.m$variable[i] == 'IGF2') sm_exp.t.m$gene_use[i] = 'fetal'
}

sm_exp.t.m$gene_use = factor(sm_exp.t.m$gene_use, levels = c('fetal', 'adult'))
sm_exp.t.m$cell_types = factor(sm_exp.t.m$cell_types, levels = c("Fetal cardiac smooth muscle cells", "Fetal other smooth muscle cells", "Adult other smooth muscle cells", "Smooth_muscle_teratoma"))

aggregate(value ~ cell_types + variable, sm_exp.t.m, length) #add data points for GCT data for IGF2

p3 = ggplot(sm_exp.t.m, mapping = aes(x = gene_use, y = value)) + 
  geom_quasirandom(dodge.width=.8, cex=1, aes(x = gene_use, y = value), shape = 21, fill = 'transparent', col = 'transparent') + 
  geom_beeswarm(data = sm_exp.t.m[sm_exp.t.m$cell_types == 'Smooth_muscle_teratoma' & sm_exp.t.m$variable == 'IGF2', ], mapping = aes(x = gene_use, y = value), dodge.width=1, cex=1, shape = 21, fill = 'black', col = 'black') + 
  facet_wrap(. ~ cell_types, strip.position = 'bottom', ncol = 8) +
  theme_pubr() + 
  theme(axis.text.x = element_blank(), 
        panel.spacing = unit(0, "mm"), 
        strip.background = element_blank(), 
        strip.placement = "outside", 
        legend.position = 'none', 
        axis.ticks.x = element_blank()) +
  labs(y = 'Log2(TPM) expression', x = '') + 
  geom_pointrange(mapping = aes(col = gene_use), stat = "summary", shape=19, fun.min = function(z) {quantile(z,0.25)}, 
                  fun.max = function(z) {quantile(z,0.75)}, 
                  fun = median, size = 0.8, fatten = 5, linetype = 2) + 
  scale_color_manual(values = c('#FF6C5C', '#58CCED'))

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/log2tpm_expression_gct_vs_ref_tissue20210305.pdf', width = 14, height = 9, useDingbats = F)
ggdraw() +
  draw_plot(p1, 0, 2/3, 8/11, 1/3) +
  draw_plot(p2, 0, 1/3, 1, 1/3) +
  draw_plot(p3, 0, 0, 4.25/11, 1/3) +
  draw_plot_label(c("A", "B", "C"), c(0, 0, 0), c(1, 2/3, 1/3), size = 15)
dev.off()

```

#Volcano plots of differential expression analysis between embryonal carcinoma and other GCT tissues
```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de')

library(EnhancedVolcano)
library(cowplot)

#read in example DE analyses
ne_v_ec = read.table('neuroepi_ec_voom_de_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)
sm_v_ec = read.table('smoothmusc_ec_voom_de_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)
ct_v_ec = read.table('cartilage_ec_voom_de_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)

p1 = EnhancedVolcano(ne_v_ec,
    lab = rownames(ne_v_ec),
    x = 'logFC',
    y = 'adj.P.Val', selectLab = c('SOX1', 'SOX3', 'PAX6', 'CDH2', 'SOX2', 'POU5F1', 'NANOG'), 
    title = 'Neuroepithelium',
    subtitle = '',
    labFace = 'bold', 
    boxedLabels = T, 
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    typeConnectors = 'open',
    colConnectors = 'black', 
    legendPosition = 'none', 
    pCutoff = 0.01,
    FCcutoff = 2,
    labSize = 5, 
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    caption = '')
p2 = EnhancedVolcano(sm_v_ec,
    lab = rownames(sm_v_ec),
    x = 'logFC',
    y = 'adj.P.Val', selectLab = c('MYOCD', 'ACTA2', 'MIR145', 'MIR143HG', 'SOX2', 'POU5F1', 'NANOG'), 
    title = 'Smooth muscle teratoma',
    subtitle = '',
    labFace = 'bold', 
    boxedLabels = T, 
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    typeConnectors = 'closed',
    colConnectors = 'black', 
    legendPosition = 'none', 
    pCutoff = 0.01,
    FCcutoff = 2,
    labSize = 5, 
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    caption = '')
p3 = EnhancedVolcano(ct_v_ec,
    lab = rownames(ct_v_ec),
    x = 'logFC',
    y = 'adj.P.Val', selectLab = c('COL2A1', 'SOX9', 'ACAN', 'COMP', 'SOX2', 'POU5F1', 'NANOG'), 
    title = 'Cartilage teratoma',
    subtitle = '',
    labFace = 'bold', 
    boxedLabels = T, 
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    typeConnectors = 'closed',
    colConnectors = 'black', 
    legendPosition = 'none', 
    pCutoff = 0.01,
    FCcutoff = 2,
    labSize = 5,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    caption = '')

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_tissues_v_ec_de_plot_20210304.pdf', height = 6, width = 18, useDingbats = F)
plot_grid(p3, p1, p2, ncol = 3, align = 'h')
dev.off()

```

#Cytoband-level gene expression by GCT tissue

```{r}

library(msigdbr)
library(limma)
library(scales)
library(ggpubr)
library(reshape2)
library(ComplexHeatmap)
library(circlize)

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de')

gct_v = readRDS('limma_voom_de_LCM_GCT_DE_voom_object_20210304.RDS')
gct_fit = readRDS('limma_voom_de_LCM_GCT_DE_model_fit_20210304.RDS')

contr.matrix <- makeContrasts(
  fat_v_st = Adipose_teratoma - Seminiferous_tubule,
  dys_v_st = Dysgerminoma - Seminiferous_tubule,
  ec_v_st = Embryonal_carcinoma - Seminiferous_tubule,
  gcnis_v_st = GCNIS - Seminiferous_tubule,
  cartilage_v_st = Immature_cartilage_teratoma - Seminiferous_tubule, 
  glandular_v_st = Mature_glandular_teratoma - Seminiferous_tubule, 
  ne_v_st = Neuroepithelium - Seminiferous_tubule, 
  other_ter_v_st = Other_epithelial_teratoma - Seminiferous_tubule, 
  muscle_v_st = Smooth_muscle_teratoma - Seminiferous_tubule, 
  sct_v_st = Syncytiotrophoblasts - Seminiferous_tubule, 
  yst_v_st = Yolk_sac_tumour - Seminiferous_tubule, 
  sem_v_st = Seminoma - Seminiferous_tubule,
  str_v_st = Stroma - Seminiferous_tubule,
  levels = colnames(gct_fit$design))

all_hum_gene_sets = msigdbr(species = "Homo sapiens")

c1_gene_set = list()

for(i in unique(all_hum_gene_sets[all_hum_gene_sets$gs_cat == 'C1', ]$gs_name)){
  c1_gene_set[[i]] = all_hum_gene_sets[all_hum_gene_sets$gs_cat == 'C1' & all_hum_gene_sets$gs_name == i, ]$human_gene_symbol
}
c1_idx <- ids2indices(c1_gene_set,id=rownames(gct_v))

all_histo_regions_enrich = c()
for(i in 1:ncol(contr.matrix)){
  temp = camera(gct_v, c1_idx, gct_fit$design, contrast = contr.matrix[,i])
  tissue = names(contr.matrix[,i][contr.matrix[,i] == 1])
  temp$tissue = tissue
  temp$cytoband = row.names(temp)
  row.names(temp) = NULL
  all_histo_regions_enrich = rbind(all_histo_regions_enrich, temp)
}

all_histo_regions_enrich$chr = substr(unlist(lapply(strsplit(all_histo_regions_enrich$cytoband, 'p|q'), '[[', 1)), 4, 5)
all_histo_regions_enrich$chr = factor(all_histo_regions_enrich$chr, levels = c(1:22, 'X', 'Y'))
all_histo_regions_enrich$arm = substr(unlist(lapply(strsplit(all_histo_regions_enrich$cytoband, paste(unique(paste0('chr', all_histo_regions_enrich$chr)), collapse = '|')), '[[', 2)), 0, 1)
all_histo_regions_enrich$cytoband_pos = substr(unlist(lapply(strsplit(all_histo_regions_enrich$cytoband, 'p|q'), '[[', 2)), 0, 2)
  
all_histo_regions_enrich = all_histo_regions_enrich[order(all_histo_regions_enrich$tissue, all_histo_regions_enrich$chr, all_histo_regions_enrich$arm, all_histo_regions_enrich$cytoband_pos),]
all_histo_regions_enrich$cytoband = factor(all_histo_regions_enrich$cytoband, levels = unique(all_histo_regions_enrich$cytoband))
all_histo_regions_enrich$sig_variable = 1 / all_histo_regions_enrich$FDR #invert for plotting purposes

for(i in 1:nrow(all_histo_regions_enrich)){
  if(all_histo_regions_enrich$Direction[i] == 'Down') all_histo_regions_enrich$sig_variable[i] = -1 * all_histo_regions_enrich$sig_variable[i]
}

write.table(all_histo_regions_enrich, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/gct_cytoband_region_enrichment20210305.txt', col.names = T, row.names = F, quote = F, sep = '\t')

all_histo_regions_enrich_wide <- dcast(all_histo_regions_enrich, tissue ~ cytoband, value.var="sig_variable")
row.names(all_histo_regions_enrich_wide) = all_histo_regions_enrich_wide$tissue

all_histo_regions_enrich_wide = all_histo_regions_enrich_wide[,2:297]
all_histo_regions_enrich_wide_auto = all_histo_regions_enrich_wide[, !grepl('chrX|chrY', colnames(all_histo_regions_enrich_wide)),]
 
col_fun = colorRamp2(c(-100, -10, 0, 10, 100), c('purple', '#520160', 'black', '#435E1C', '#95D840'))
chrom_order = factor(unlist(lapply(strsplit(colnames(all_histo_regions_enrich_wide_auto), 'p|q'), '[[', 1)), levels = paste0('chr', 1:22))
levels(chrom_order) = 1:22

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/enriched_regions_gct_expression20210305.pdf', width = 15, height = 8)
Heatmap(all_histo_regions_enrich_wide_auto, cluster_columns = F, cluster_rows = F, col = col_fun, column_split = chrom_order, heatmap_legend_param = list(title = "FDR", at = c(-100, -10, 10, 100), labels = c(0.01, 0.1, 0.1, 0.01), legend_height = unit(4, "cm")), show_column_names = F, row_names_side = 'left', row_order = c('GCNIS', 'Dysgerminoma', 'Seminoma', 'Embryonal_carcinoma', 'Yolk_sac_tumour', 'Syncytiotrophoblasts', 'Neuroepithelium', 'Mature_glandular_teratoma', 'Other_epithelial_teratoma', 'Adipose_teratoma', 'Smooth_muscle_teratoma', 'Immature_cartilage_teratoma', 'Stroma'))
dev.off()

```

#Which 12p genes are upregulated in which GCT tissues?

```{r}

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de')

invasive_comparison = list.files(pattern = '_normtestis_voom_de_20210304.txt')[!grepl('gcnis|invasive_gct', list.files(pattern = '_normtestis_voom_de_20210304.txt'))]

raw_data = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/filtering/germ_cell_lcm_rna_preQC.txt', header = T, sep = '\t', stringsAsFactors = F) #raw count data
row.names(raw_data) = raw_data$Geneid
raw_data$Start = unlist(lapply(strsplit(raw_data$Start, ';'), '[[', 1))
raw_data$Start = as.numeric(raw_data$Start)
genes_list = row.names(raw_data[raw_data$Chr == 12 & raw_data$Start < 35800000,]) #759


all_tissue_df = c()
for(i in invasive_comparison){
  tissue_file = read.table(i, header = T, sep = '\t', stringsAsFactors = F)
  sig_up = tissue_file[tissue_file$logFC >= 2 & tissue_file$adj.P.Val < 0.01,]
  sig_up = sig_up[row.names(sig_up) %in% genes_list,]
  tissue_df = data.frame(row.names(sig_up), i)
  all_tissue_df = rbind(all_tissue_df, tissue_df)
}

all_tissue_df$i = as.character(all_tissue_df$i)
all_tissue_df$i = unlist(strsplit(all_tissue_df$i, '_normtestis_voom_de_20210304.txt'))
all_tissue_tab = as.data.frame(table(all_tissue_df$row.names.sig_up.))
all_tissue_tab = all_tissue_tab[order(all_tissue_tab$Freq, decreasing = T),]

tf_list = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/reference_datasets/church_tfs.txt', header = F, sep = '\t', stringsAsFactors = F)[,1]

all_tissue_tab$Var1 = as.character(all_tissue_tab$Var1)
all_tissue_tab$is_tf = 'no'
for(i in 1:nrow(all_tissue_tab)){
  if(all_tissue_tab$Var1[i] %in% tf_list) all_tissue_tab$is_tf[i] = 'yes'
}

names(all_tissue_tab) = c('12p gene', 'No. of tissues upregulated in (/12)', 'Transcription factor')

all_tissue_tab$tissues_involved = NA
for(j in 1:nrow(all_tissue_tab)){
  all_tissue_tab$tissues_involved[j] = paste(all_tissue_df[all_tissue_df$row.names.sig_up. == all_tissue_tab$`12p gene`[j],]$i, collapse = ', ')
}

write.table(all_tissue_tab, '/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/enriched_12p_genes/gct_tissues_enriched_for_12p_genes_20210305.txt', col.names = T, row.names = F, sep = '\t', quote = F)

```


#Limma-Voom DE analysis between different all invasive GCT tissues together and normal testis for chromosome 12p composite plot


```{r}

library(limma)
library(edgeR)

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing')

raw_data = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/filtering/germ_cell_lcm_rna_preQC.txt', header = T, sep = '\t', stringsAsFactors = F) #raw count data
metadata = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_RNA_metadata.txt', header = T, sep = '\t', stringsAsFactors = F) #read in metadata for filtered, final cuts
row.names(metadata) = metadata$Sample_ID
gct_mat = raw_data[, 7:ncol(raw_data)]
row.names(gct_mat) = raw_data$Geneid

#remove cuts that failed QC
gct_mat = gct_mat[, colnames(gct_mat) %in% metadata[metadata$Updated.Description != 'GCNIS',]$Sample_ID] #382 samples after removing GCNIS

#define factors that will be input into the model
histology = factor(metadata[colnames(gct_mat), ]$Updated.Description)
tumour = factor(metadata[colnames(gct_mat), ]$Normal.Tumour)
metadata = metadata[colnames(gct_mat),]

#preprocessing
y <- DGEList(counts = gct_mat, group = tumour)
keep <- filterByExpr(y, group = tumour)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y) #12431 genes kept
y <- estimateCommonDisp(y) #to generate pseudocounts, adjusted for library size

#visualise if microbiopsies from the same histology cluster and if there is patient effect
plotMDS(y, labels = tumour, col = as.numeric(histology))

#specify model to be fitted, no intercept to be fitted, each normal-tumour status will have its own group mean
design = model.matrix(~ 0 + tumour)
colnames(design) = levels(tumour)

#run double voom to optimise regression weights, adjust for histology as 'block' to find underlying generic tumour signal
v <- voom(y, design)
corfit <- duplicateCorrelation(v, design, block = histology)
v <- voom(y, design, block = histology, correlation = corfit$consensus)
corfit <- duplicateCorrelation(v, design, block = histology)
fit <- lmFit(v, design, block = histology, correlation = corfit$consensus)
corfit$consensus #0.1980107

setwd('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de')

saveRDS(v, 'LCM_invasive_GCT_v_testis_DE_voom_object_20210305.RDS')
saveRDS(fit, 'LCM_invasive_GCT_v_testis_DE_model_fit_20210305.RDS')

v = readRDS('LCM_invasive_GCT_v_testis_DE_voom_object_20210305.RDS')
fit = readRDS('LCM_invasive_GCT_v_testis_DE_model_fit_20210305.RDS')

cont_matrix = makeContrasts(Tumour - Normal, levels = colnames(coef(fit)))
fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2)
top.table <- topTable(fit2, number = nrow(fit$coefficients), sort.by = "P")

raw_data = read.table("/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/filtering/germ_cell_lcm_rna_preQC.txt", header = T, sep = '\t', stringsAsFactors = F, check.names = T)
row.names(raw_data) = raw_data$Geneid
raw_data$Start = unlist(lapply(strsplit(raw_data$Start, ';'), '[[', 1))

top.table$chr = raw_data[row.names(top.table),]$Chr
top.table$start = raw_data[row.names(top.table),]$Start
top.table$start = as.numeric(top.table$start)

write.table(top.table, 'invasive_gct_v_normtestis_voom_de_20210304.txt', col.names = T, sep = '\t', row.names = T, quote = F)

```

##Plot

```{r}

#fetch cn data from Fig. 1

cn_df = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/lcm_gct_cn_10000_bp_bin_per_GCT_WG_median_extra_cols.txt', header = T, sep = '\t', stringsAsFactors = F)

#limit to chromosome 12 (cumulative coordinates from 10kb bins in file)

lcm_chr12_df = cn_df[(cn_df$start > 1950970000) & (cn_df$end <= 2084830000), ]
lcm_chr12_df$chr = 'chr12'
lcm_chr12_df$start = lcm_chr12_df$start - 1950970000
lcm_chr12_df$end = lcm_chr12_df$end - 1950970000
lcm_chr12_df = lcm_chr12_df[!is.na(lcm_chr12_df$all_samples),]

#fetch logFC data from invasive v normal comparison

top.table.lcm = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de/invasive_gct_v_normtestis_voom_de_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)

#limit to chromsome 12 data for plotting
lcm_chr12_exp_data = top.table.lcm[top.table.lcm$chr == 12,]
lcm_chr12_exp_data$chr = paste0('chr', lcm_chr12_exp_data$chr)
lcm_chr12_exp_data = lcm_chr12_exp_data[order(lcm_chr12_exp_data$start),]

library(karyoploteR)
library(zoo)

#generic plotting settings
chromosomes = 'chr12'
dna_r0 <- 0.9
dna_r1 <- 0
rna_r0 <- 0
rna_r1 <- 0.9
karyoplotr_plottype = 1

#in-house rna.data
window_size = 50 #must be an even number
y_data = rollmean(lcm_chr12_exp_data$logFC, k = window_size, align = 'center')
x_data = lcm_chr12_exp_data$start[(window_size / 2):(length(lcm_chr12_exp_data$start) - (window_size / 2))]
pos_y_data = ifelse(y_data > 0, y_data, 0) 
neg_y_data = ifelse(y_data < 0, y_data, 0)

##plot in-house data
title = 'In-house'

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_cn_v_exp_karyoplotter_20210305.pdf', height = 6, width = 6, useDingbats = F)
kp <- karyoploteR::plotKaryotype(genome = "hg19", chromosomes = 'chr12', plot.type = karyoplotr_plottype, cex = 0.85, main = gsub(".*__", "", title))
karyoploteR::kpDataBackground(kp, r0 = dna_r0, r1 = dna_r1, col = 'grey90')
karyoploteR::kpAxis(karyoplot = kp, col = "black", ymin = -1.5, ymax = 1.5, r0 = rna_r0, r1 = rna_r1, cex = 1, tick.pos = c(-1.5, -1, -0.5, 0, 0.5, 1, 1.5), side = 2)
karyoploteR::kpBars(kp, chr = 'chr12', x0 = x_data - 100000, x1 = x_data + 100000, y0 = 0, y1 = pos_y_data, col = '#95D840', data.panel = 1, ymin = -1.5, ymax = 1.5, r0 = rna_r0, r1 = rna_r1, clipping = T, border = 'NA')
karyoploteR::kpBars(kp, chr = 'chr12', x0 = x_data - 100000, x1 = x_data + 100000, y0 = 0, y1 = neg_y_data, col = '#A020F0', data.panel = 1, ymin = -1.5, ymax = 1.5, r0 = rna_r0, r1 = rna_r1, clipping = T, border = 'NA')
karyoploteR::kpAxis(karyoplot = kp, col = "black", ymin = 6, ymax = -3, r1 = dna_r1, r0 = dna_r0, cex = 1, tick.pos = c(-3, 0, 3, 6), side = 1)
karyoploteR::kpPoints(kp, chr = lcm_chr12_df$chr, x = lcm_chr12_df$start, y = lcm_chr12_df$all_samples, col = "black", ymin = 6, ymax = -3, r0 = dna_r0, r1 = dna_r1, clipping = T) #reintroduced prepubertal samples
dev.off()
```

#12p gene enrichment compared to regions of near baseline ploidy

```{r}

#fetch cn data from Fig. 1

cn_df = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/combined_CN_RNA_exp/lcm_gct_cn_10000_bp_bin_per_GCT_WG_median_extra_cols.txt', header = T, sep = '\t', stringsAsFactors = F)

#limit to chromosome 12 (cumulative coordinates from 10kb bins in file)

lcm_chr12_df = cn_df[(cn_df$start > 1950970000) & (cn_df$end <= 2084830000), ]
lcm_chr12_df$chr = 'chr12'
lcm_chr12_df$start = lcm_chr12_df$start - 1950970000
lcm_chr12_df$end = lcm_chr12_df$end - 1950970000
lcm_chr12_df = lcm_chr12_df[!is.na(lcm_chr12_df$all_samples),]

#fetch logFC data from invasive v normal comparison

top.table.lcm = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/limma_voom_de/invasive_gct_v_normtestis_voom_de_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)
top.table.lcm = top.table.lcm[top.table.lcm$chr %in% c(1:22),] #only performed copy number plotting across autosomal genome for comparison

#limit to chromsome 12 data for plotting
lcm_chr12_exp_data = top.table.lcm[top.table.lcm$chr == 12,]
lcm_chr12_exp_data$chr = paste0('chr', lcm_chr12_exp_data$chr)
lcm_chr12_exp_data = lcm_chr12_exp_data[order(lcm_chr12_exp_data$start),]


#get regions with a CN-ploidy within 0.5 of the overall ploidy located outside of chr12
av_cn = cn_df[!(is.na(cn_df$all_samples)) & (cn_df$all_samples > -0.5) & (cn_df$all_samples < 0.5) & ((cn_df$start < 1950970000) | (cn_df$start > 2084830000)),]

#provided cumulative loci for genes
top.table.lcm$mod_start = top.table.lcm$start
for(k in 1:nrow(top.table.lcm)) {
    if(top.table.lcm$chr[k] != "1"){
        top.table.lcm$mod_start[k] = top.table.lcm$start[k] + chrom_sizes$prior_wg_length[which(chrom_sizes$chr == top.table.lcm$chr[k])]
      }
}

top.table.lcm$av_cn_region = NA

for(i in 1:nrow(top.table.lcm)){
  if(nrow(av_cn[(av_cn$start < top.table.lcm$mod_start[i]) & (av_cn$end > top.table.lcm$mod_start[i]), ]) > 0){
    top.table.lcm$av_cn_region[i] = 'Yes'
  }
  if(nrow(av_cn[(av_cn$start < top.table.lcm$mod_start[i]) & (av_cn$end > top.table.lcm$mod_start[i]), ]) == 0){
    top.table.lcm$av_cn_region[i] = 'No'
  }
}

n_12p_genes = nrow(top.table.lcm[top.table.lcm$chr == 12 & top.table.lcm$start < 35800000,]) #226 used
mean_12p = mean(top.table.lcm[top.table.lcm$chr == 12 & top.table.lcm$start < 35800000,]$logFC) # 0.686984

nRand = 100000
randScores=c()
set.seed(42)
av_cn_exp_regions = top.table.lcm[top.table.lcm$av_cn_region == 'Yes',]

for(i in 1:nRand){
  randScores=c(randScores, mean(sample(av_cn_exp_regions$logFC, n_12p_genes, replace = F)))
}

pVal = sum(randScores >= mean_12p)/nRand #0...

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/rand_sampling_gene_exp_v_12p_20210305.pdf', height = 3, width = 4)
ggplot() + geom_density(mapping = aes(x = randScores), fill = '#999999') + xlim(-2, 1) + theme_pubr() + geom_vline(xintercept = mean_12p) + labs(x = 'log2FC of gene expression in regions with baseline ploidy') + coord_cartesian(expand = F)
dev.off()

```

#Sample proximity scores
##Genomic

```{r}

library(ggplot2)
library(ggpubr)
library(ggbeeswarm)

dna_cuts = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_DNA_supplementary_data.txt', header = T, sep = '\t', stringsAsFactors = F)

mixed_feature_cuts = dna_cuts[dna_cuts$Histology != 'GCNIS' & dna_cuts$Diagnosis == 'Post-pubertal mixed germ cell tumour',]

pluri_mixed_feature_cuts = mixed_feature_cuts[!mixed_feature_cuts$Case_ID %in% c('PD45545', 'PD46966', 'PD46969'),] #only one histology in these tumours represented or multiple but spread a thinly across multiple biopsies unevenly which would distort a genetic proximity score

#between histologies
interhisto_g_list = list()
for(i in unique(pluri_mixed_feature_cuts$Case_ID)){ #per patient
  assessed_histo = c() #record of histologies already analysed
  g_prox_score_vec = c() #vector of per patient comparison scores
  for(j in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i,]$Histology)){ #per histology
    assessed_histo = c(assessed_histo, j) #add histology to list not to compare against
    if(j != unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i,]$Histology)[length(unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i,]$Histology))]){ #no groups left for the final histology to be compared against
        for(k in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j,]$Sample)){ #each patient sample per histology
      choice_sample = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/SNV/', k, '_post_swater_final_snvs.txt'), header = T, sep = '\t', stringsAsFactors = F)
      choice_sample = choice_sample[choice_sample$VAF >= 0.1,] #remove low level contamination from other histologies
      for(l in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & !(pluri_mixed_feature_cuts$Histology %in% assessed_histo),]$Sample)){
        comparator_sample = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/SNV/', l, '_post_swater_final_snvs.txt'), header = T, sep = '\t', stringsAsFactors = F)
        comparator_sample = comparator_sample[comparator_sample$VAF >= 0.1,] #remove low level contamination from other histologies
        g_prox = length(intersect(choice_sample$ID, comparator_sample$ID)) / ((nrow(choice_sample) + nrow(comparator_sample)) / 2) #average sharedness of high VAF mutations
        g_prox_score_vec = c(g_prox_score_vec, g_prox) #add pairwise comparison to patient vector
      }
    }
    }
  }
  interhisto_g_list = c(interhisto_g_list, list(g_prox_score_vec))
}

names(interhisto_g_list) = unique(pluri_mixed_feature_cuts$Case_ID)
sapply(interhisto_g_list, median)

#within histologies
intrahisto_g_list = list()
for(i in unique(pluri_mixed_feature_cuts$Case_ID)){ #per patient
  g_prox_score_vec = c() #vector of per patient comparison scores
  for(j in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i,]$Histology)){ #per histology
    if(nrow(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j,]) > 1){ #if histology has more than one microbiopsy
      assessed_samples = c() #record of samples already analysed
      for(k in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j,]$Sample)){ #per sample
        assessed_samples = c(assessed_samples, k) #add to vector of analysed samples
        choice_sample = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/SNV/', k, '_post_swater_final_snvs.txt'), header = T, sep = '\t', stringsAsFactors = F)
        choice_sample = choice_sample[choice_sample$VAF >= 0.1,] #remove low level contamination from other histologies
        if(k != unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j,]$Sample)[length(unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j,]$Sample))]){ #as the last sample cannot be compared against itself
          for(l in unique(pluri_mixed_feature_cuts[pluri_mixed_feature_cuts$Case_ID == i & pluri_mixed_feature_cuts$Histology == j & !(pluri_mixed_feature_cuts$Sample %in% assessed_samples),]$Sample)){
          comparator_sample = read.table(paste0('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/final_variant_calls/SNV/', l, '_post_swater_final_snvs.txt'), header = T, sep = '\t', stringsAsFactors = F)
          comparator_sample = comparator_sample[comparator_sample$VAF >= 0.1,] #remove low level contamination from other histologies
        g_prox = length(intersect(choice_sample$ID, comparator_sample$ID)) / ((nrow(choice_sample) + nrow(comparator_sample)) / 2) #average sharedness of high VAF mutations
        g_prox_score_vec = c(g_prox_score_vec, g_prox) #add pairwise comparison to patient vector
        }
        }
        
      }
    }
  }
  intrahisto_g_list = c(intrahisto_g_list, list(g_prox_score_vec))
}

names(intrahisto_g_list) = unique(pluri_mixed_feature_cuts$Case_ID)
sapply(intrahisto_g_list, median)

#combine results
interhisto_g_df = data.frame(unlist(interhisto_g_list))
interhisto_g_df$patient = substr(row.names(interhisto_g_df), 0, 7)
interhisto_g_df$comparison = 'Inter-histology'
row.names(interhisto_g_df) = NULL
colnames(interhisto_g_df) = c('gen_prox', 'patient', 'comparison')

intrahisto_g_df = data.frame(unlist(intrahisto_g_list))
intrahisto_g_df$patient = substr(row.names(intrahisto_g_df), 0, 7)
intrahisto_g_df$comparison = 'Intra-histology'
row.names(intrahisto_g_df) = NULL
colnames(intrahisto_g_df) = c('gen_prox', 'patient', 'comparison')

all_histo_g_df = rbind(interhisto_g_df, intrahisto_g_df)
all_histo_g_df$comparison = factor(all_histo_g_df$comparison, levels = c('Intra-histology', 'Inter-histology'))
all_histo_g_df$patient = as.factor(all_histo_g_df$patient)

#are these differences significant? no
PD43296_g = wilcox.test(gen_prox ~ comparison, all_histo_g_df[all_histo_g_df$patient == 'PD43296',]) #W = 24522, p-value = 0.1284
PD43298_g = wilcox.test(gen_prox ~ comparison, all_histo_g_df[all_histo_g_df$patient == 'PD43298',]) #W = 12651, p-value = 0.8893
PD45543_g = wilcox.test(gen_prox ~ comparison, all_histo_g_df[all_histo_g_df$patient == 'PD45543',]) #W = 4, p-value = 1
PD45544_g = wilcox.test(gen_prox ~ comparison, all_histo_g_df[all_histo_g_df$patient == 'PD45544',]) #W = 151, p-value = 0.4753
PD46269_g = wilcox.test(gen_prox ~ comparison, all_histo_g_df[all_histo_g_df$patient == 'PD46269',]) #W = 145, p-value = 0.6483

#adjust for multiple testing
p_g_vec = c(PD43296_g$p.value, PD43298_g$p.value, PD45543_g$p.value, PD45544_g$p.value, PD46269_g$p.value) 
p.adj_g_vec = p.adjust(p_g_vec, method="BH")
names(p.adj_g_vec) = unique(pluri_mixed_feature_cuts$Case_ID)

ggplot(all_histo_g_df) + 
  geom_quasirandom(mapping = aes(x = gen_prox, y = comparison, col = comparison), width = 0.3, groupOnX = F) +
  geom_pointrange(mapping = aes(x = gen_prox, y = comparison), stat = "summary", 
                  shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill="black") +
  facet_grid(patient ~ ., scales = 'free_y', space = 'free_y') + theme_pubr() + theme(axis.text.y = element_blank(), panel.spacing = unit(0, 'mm'), strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = 11)) + xlab('Genetic similarity') + ylab ('') + scale_y_discrete(position = "right")  + scale_x_continuous(expand = c(0,0), limits = c(0,1))

p1 = ggplot(all_histo_g_df) + 
  geom_boxplot(mapping = aes(x = gen_prox, y = comparison, fill = comparison), width = 0.5) +
  facet_grid(patient ~ ., scales = 'free_y', space = 'free_y') + theme_pubr() + theme(axis.text.y = element_blank(), panel.spacing = unit(0, 'mm'), strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = 11)) + xlab('Genetic similarity') + ylab ('') + scale_y_discrete(position = "right")  + scale_x_continuous(expand = c(0,0), limits = c(0,1)) + scale_fill_manual(values = c('#F8766D', '#00BFC4'), name = 'Comparison')

```

##Transcriptomic

```{r}

#libraries to use
#library(lsa)
library(Seurat)
library(ggplot2)
library(ggpubr)
library(ggbeeswarm)

#genes to remove to leave ones of more biological interest to compare
source('/lustre/scratch119/casm/team294rr/to3/testes/tumour/RNA_analysis/geneSets.R')
prot_cod = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/RNA_analysis/protein_coding_genes.txt', header = F, sep='\t', stringsAsFactors = F)[,1]

#read in data
tpm = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/analyses/RNA/pre_processing/tmm_norm_gct_tpm_20210304.txt', header = T, sep = '\t', stringsAsFactors = F)
log2tpm <- log2(as.matrix(tpm) + 1)
log2tpm = log2tpm[(row.names(log2tpm) %in% prot_cod) & !(row.names(log2tpm) %in% unique(c(hgGenes, hkGenes, igGenes, riboGenes, riboRNAGenes, cc.genes.updated.2019$s.genes, cc.genes.updated.2019$g2m.genes))),]
metadata = read.table('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/supplementary_data/GCT_RNA_metadata.txt', header = T, sep = '\t', stringsAsFactors = F)
row.names(metadata) = metadata$Sample_ID
inv_mixed_rna = metadata[metadata$Diagnosis == 'Post-pubertal mixed germ cell tumour' & metadata$Normal.Tumour == 'Tumour' & metadata$Updated.Description != 'GCNIS',]
inv_mixed_rna = inv_mixed_rna[inv_mixed_rna$Case_ID %in% c('PD43296', 'PD43298', 'PD45543', 'PD45544', 'PD46269'),] #make direct contrast in each tumour to their transcriptome

#between histologies
interhisto_t_list = list()
for(i in unique(inv_mixed_rna$Case_ID)){ #per patient
  assessed_histo = c() #record of histologies already analysed
  t_prox_score_vec = c() #vector per patient comparison scores
  for(j in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i,]$Updated.Description)){
    assessed_histo = c(assessed_histo, j) #add histology to list to not compare against
    if(length(assessed_histo) != length(unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i,]$Updated.Description))){ #stop when you are at the last histology as you have already compared it to all of the others
       for(k in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j,]$Sample_ID)){ #samples per histology per tumour
      choice_sample = log2tpm[, k] #vector from sample in histology of interest
      for(l in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & !(inv_mixed_rna$Updated.Description %in% assessed_histo),]$Sample_ID)){
        comparator_sample = log2tpm[, l] #vector from a sample in another histology
        t_prox = cor(choice_sample, comparator_sample, method = 'pearson')
        t_prox_score_vec = c(t_prox_score_vec, t_prox)
    }
    }
    }
   
  }
  interhisto_t_list = c(interhisto_t_list, list(t_prox_score_vec))
}

names(interhisto_t_list) = unique(inv_mixed_rna$Case_ID)

sapply(interhisto_t_list, median)

#within histologies
intrahisto_t_list = list()
for(i in unique(inv_mixed_rna$Case_ID)){ #per patient
  t_prox_score_vec = c() #vector per patient comparison scores
  for(j in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i,]$Updated.Description)){ #per histology
      if(nrow(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j,]) > 1){ #if histology has more than one microbiopsy
      assessed_samples = c() #record of samples already analysed
      for(k in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j,]$Sample)){ #per sample
        assessed_samples = c(assessed_samples, k) #add to vector analysed
        choice_sample = log2tpm[, k]
        
        if(k != unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j,]$Sample)[length(unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j,]$Sample))]){ #so the last sample isn't compared against itself
          for(l in unique(inv_mixed_rna[inv_mixed_rna$Case_ID == i & inv_mixed_rna$Updated.Description == j & !(inv_mixed_rna$Sample %in% assessed_samples),]$Sample)){
          comparator_sample = log2tpm[, l]
          t_prox = cor(choice_sample, comparator_sample, method = 'pearson')
          t_prox_score_vec = c(t_prox_score_vec, t_prox)
        }
        }
      }
    }
  }
  intrahisto_t_list = c(intrahisto_t_list, list(t_prox_score_vec))
}

names(intrahisto_t_list) = unique(inv_mixed_rna$Case_ID)
sapply(intrahisto_t_list, median)

#control - against normal testis
norm_control = metadata[metadata$Normal.Tumour == 'Normal',]

norm_tumour_t_list = list()
for(i in unique(inv_mixed_rna$Case_ID)){ #per patient
  t_prox_score_vec = c() #vector per patient comparison scores
  for(j in inv_mixed_rna[inv_mixed_rna$Case_ID == i,]$Sample_ID){
    choice_sample = log2tpm[, j]
    for(k in norm_control$Sample_ID){
      comparator_sample = log2tpm[, k]
      t_prox = cor(choice_sample, comparator_sample, method = 'pearson')
      #t_prox = cosine(choice_sample, comparator_sample)[1,1]
      #t_prox = cor(choice_sample, comparator_sample, method = 'spearman')
      t_prox_score_vec = c(t_prox_score_vec, t_prox)
    }
  }
  norm_tumour_t_list = c(norm_tumour_t_list, list(t_prox_score_vec))
}

names(norm_tumour_t_list) = unique(inv_mixed_rna$Case_ID)

sapply(norm_tumour_t_list, median)

#add together
interhisto_t_df = data.frame(unlist(interhisto_t_list))
interhisto_t_df$patient = substr(row.names(interhisto_t_df), 0, 7)
interhisto_t_df$comparison = 'Inter-histology'
row.names(interhisto_t_df) = NULL
colnames(interhisto_t_df) = c('t_prox', 'patient', 'comparison')

intrahisto_t_df = data.frame(unlist(intrahisto_t_list))
intrahisto_t_df$patient = substr(row.names(intrahisto_t_df), 0, 7)
intrahisto_t_df$comparison = 'Intra-histology'
row.names(intrahisto_t_df) = NULL
colnames(intrahisto_t_df) = c('t_prox', 'patient', 'comparison')

norm_tumour_t_df = data.frame(unlist(norm_tumour_t_list))
norm_tumour_t_df$patient = substr(row.names(norm_tumour_t_df), 0, 7)
norm_tumour_t_df$comparison = 'Normal v Tumour'
row.names(norm_tumour_t_df) = NULL
colnames(norm_tumour_t_df) = c('t_prox', 'patient', 'comparison')

all_histo_t_df = rbind(interhisto_t_df, intrahisto_t_df, norm_tumour_t_df)

ggplot(all_histo_t_df) + 
  geom_quasirandom(mapping = aes(x = comparison, y = t_prox, col = comparison), width = 0.3) +
  geom_pointrange(mapping = aes(x = comparison, y = t_prox), stat = "summary", 
                  shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill="black") +
  facet_grid(. ~ patient, scales = 'free_x', space = 'free_x', switch = 'x') + theme_pubr() + theme(axis.text.x = element_blank(), panel.spacing = unit(0, 'mm'), strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = 11)) + ylab('Transcriptomic similarity') + xlab ('') + ylim(0, 1)

PD43296_t = wilcox.test(t_prox ~ comparison, all_histo_t_df[all_histo_t_df$patient == 'PD43296' & all_histo_t_df$comparison %in% c('Inter-histology', 'Intra-histology'),]) #W = 117762, p-value < 2.2e-16
PD43298_t = wilcox.test(t_prox ~ comparison, all_histo_t_df[all_histo_t_df$patient == 'PD43298' & all_histo_t_df$comparison %in% c('Inter-histology', 'Intra-histology'),]) #W = 51363, p-value < 2.2e-16
PD45543_t = wilcox.test(t_prox ~ comparison, all_histo_t_df[all_histo_t_df$patient == 'PD45543' & all_histo_t_df$comparison %in% c('Inter-histology', 'Intra-histology'),]) #W = 7, p-value = 0.01399
PD45544_t = wilcox.test(t_prox ~ comparison, all_histo_t_df[all_histo_t_df$patient == 'PD45544' & all_histo_t_df$comparison %in% c('Inter-histology', 'Intra-histology'),]) #W = 3492, p-value < 2.2e-16
PD46269_t = wilcox.test(t_prox ~ comparison, all_histo_t_df[all_histo_t_df$patient == 'PD46269' & all_histo_t_df$comparison %in% c('Inter-histology', 'Intra-histology'),]) #W = 15259, p-value < 2.2e-16

p_t_vec = c(PD43296_t$p.value, PD43298_t$p.value, PD45543_t$p.value, PD45544_t$p.value, PD46269_t$p.value) 
p.adj_t_vec = p.adjust(p_t_vec, method="BH")
names(p.adj_t_vec) = unique(inv_mixed_rna$Case_ID)

all_histo_t_df$comparison = factor(all_histo_t_df$comparison, levels = c('Intra-histology', 'Inter-histology', 'Normal v Tumour'))

```

##Plot
```{r}

library(cowplot)

p1 = ggplot(all_histo_g_df) + 
  geom_quasirandom(mapping = aes(x = gen_prox, y = comparison, col = comparison), width = 0.3, groupOnX = F) +
  geom_pointrange(mapping = aes(x = gen_prox, y = comparison),
                  stat = "summary", shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill = "black") +
  facet_grid(patient ~ ., scales = 'free_y', space = 'free_y') + theme_pubr() + theme(axis.text.y = element_blank(), panel.spacing = unit(0, 'mm'), strip.background = element_blank(), strip.placement = "none", strip.text = element_text(size = 11)) + xlab('Genetic similarity') + ylab ('') + scale_y_discrete(position = "right")  + scale_x_continuous(expand = c(0,0), limits = c(0,1)) + scale_color_manual(values = c("#F2AD00", "#5BBCD6"), name = 'Comparison')

p2 = ggplot(all_histo_t_df) + 
  geom_quasirandom(mapping = aes(x = t_prox, y = comparison, col = comparison), width = 0.4, groupOnX = F) +
  geom_pointrange(mapping = aes(x = t_prox, y = comparison),
                  stat = "summary", shape=19, 
                  fun.min = function(z) {quantile(z,0.25)},
                  fun.max = function(z) {quantile(z,0.75)},
                  fun = median, fill = "black") +
  facet_grid(patient ~ ., scales = 'free_y', space = 'free_y', switch = 'y') + theme_pubr() + theme(axis.text.y = element_blank(), panel.spacing = unit(0, 'mm'), strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = 11)) + xlab('Transcriptomic similarity') + ylab ('') + scale_y_discrete(position = "left")  + scale_x_reverse(expand = c(0,0), limits = c(1,0)) + scale_color_manual(values = c("#F2AD00", "#5BBCD6", '#999999'), name = 'Comparison')

pdf('/lustre/scratch119/casm/team294rr/to3/testes/tumour/manuscript/figures/gct_genetic_proximity_pearson_rna_20210305.pdf', height = 6, width = 8, useDingbats = F)
plot_grid(p1, p2, align = 'h', ncol = 2)
dev.off()

```
